---
title: "Space, Time, and Perfect Algorithms"
author: "David J. Hunter"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    df_print: paged
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(metathis)
meta() %>%
  meta_description(
    "Slides, assignments, and code for my algorithms class."
  ) %>%
  meta_name("github-repo" = "djhunter/algorithms") %>%
  meta_viewport() %>%
  meta_social(
    title = "Space, Time, and Perfect Algorithms",
    url = "https://djhunter.github.io/algorithms/index.html",
    og_author = c("David J. Hunter")
  )
```

This site provides slides, assignments, and other resources that I use in my Space, Time, and Perfect Algorithms class (CS-120, Westmont College).

# Slides 

1. [Introduction](slides/01intro_algorithms.html) 

# Written Assignments

1. [Running Time Review](assignments/01time.html)

# Other Resources

- [Syllabus](tex/cs120s21.pdf) (PDF)

All of the pages and slides on this site were authored in RMarkdown. The source code is available on GitHub: https://github.com/djhunter/algorithms

