---
title: "Space, Time, and Perfect Algorithms"
author: "David J. Hunter"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    df_print: paged
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(metathis)
meta() %>%
  meta_description(
    "Slides, assignments, and code for my algorithms class."
  ) %>%
  meta_name("github-repo" = "djhunter/algorithms") %>%
  meta_viewport() %>%
  meta_social(
    title = "Space, Time, and Perfect Algorithms",
    url = "https://djhunter.github.io/algorithms/index.html",
    og_author = c("David J. Hunter")
  )
```

This site provides slides, assignments, and other resources that I use in my Space, Time, and Perfect Algorithms class (CS-120, Westmont College).

# Slides 

1. [Introduction: Space, Time, Perfection?](slides/01intro_algorithms.html) 
2. [Recursion](slides/02recursion.html)
3. [Divide and Conquer Intro](slides/03divideandconquer.html)
4. [Divide and Conquer: Sorting](slides/04dcsorting.html)
5. [Recursion Trees](slides/05rectrees.html)
6. [Recursion Trees, continued](slides/06morerectrees.html)
7. [Backtracking](slides/07backtracking.html)
8. [More Backtracking Algorithms](slides/08backtracking2.html)
9. [Dynamic Programming: Introduction](slides/09dynprog1.html)

# Written Assignments

1. [Running Time Review](assignments/01time.html)
2. [The Baguenaudier Puzzle](assignments/02baguenaudier.html)
3. [Another Hanoi Puzzle](assignments/03morehanoi.html)
4. [The Cavarretta Sort](assignments/04cavarrettasort.html)
5. [Practice with Recursion Trees](assignments/05rectreepractice.html)
6. [More Practice with Recursion](assignments/06morerecpractice.html)
7. [Churro Cutting](assignments/07churrocutting.html)
8. [More Backtracking Exercises](assignments/08morebacktracking.html)
9. [Fast Churro Cutting](assignments/09fastchurro.html)

# Other Resources

- [Syllabus](tex/cs120S21.pdf) (PDF)
- [Textbook](http://jeffe.cs.illinois.edu/teaching/algorithms/book/Algorithms-JeffE.pdf) (PDF)

All of the pages and slides on this site were authored in RMarkdown. The source code is available on GitHub: https://github.com/djhunter/algorithms

