---
title: "Introduction: Space, Time, Perfection?"
date: "January 12, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Algorithms

## What is an Algorithm?

From [our textbook](http://jeffe.cs.illinois.edu/teaching/algorithms/book/Algorithms-JeffE.pdf):

> An algorithm is an *explicit*, precise, *unambiguous*, mechanically-executable sequence of elementary instructions, usually intended to *accomplish* a specific purpose.

From \textit{Algorithms}, by Cormen, et. al. [CLRS]:

> An algorithm is any *well-defined* computational procedure that takes some value, or set of values, as input and *produces* some value, or set of values, as output.

## Example: Peasant Multiplication

```
PeasantMultiply(X, Y):
    prod <- 0
    x <- X
    y <- Y
    while x > 0
        if x is odd
             prod <- prod + y
        x <- floor(x/2)         # mediation
        y <- y + y              # duplation
    return prod
```

- Does it work? (correctness)
- How fast is it? (time)
- How much storage does it require? (space)

# Correctness

```
PeasantMultiply(X, Y):
    prod <- 0
    x <- X
    y <- Y
    while x > 0
        if x is odd
             prod <- prod + y
        x <- floor(x/2)         # mediation
        y <- y + y              # duplation
    return prod
```

- "Easy" to show that **`XY == prod + xy`** is an *invariant* for the while loop.
- At loop termination, **`x == 0`**, so **`prod == XY`**.
- The algorithm *correctly* computes the product of `X` and `Y`.

# Running Time

```
PeasantMultiply(X, Y):
    prod <- 0
    x <- X
    y <- Y
    while x > 0
        if x is odd
             prod <- prod + y
        x <- floor(x/2)         # mediation
        y <- y + y              # duplation
    return prod
```

How many mediation and duplation operations does this algorithm require?

- Answer the [poll question](https://forms.gle/Phu8bX9DtAZ7wQJVA).

## Asymptotic notation (review)

- **$f(n)$ is $O(g(n))$** if $f(n)$ is **less than or equal** to a constant multiple of $g(n)$ for large $n$.
- **$f(n)$ is $\Omega(g(n))$** if $f(n)$ is **greater than or equal** to a constant multiple of $g(n)$ for large $n$.
- **$f(n)$ is $\Theta(g(n))$** if $f(n)$ is **both** $O(g(n))$ and $\Omega(g(n))$. In this case, we say that $f$ and $g$ are *asymptotically* the same.

We use big-$O$ to represent asymptotic upper bounds, and big-$\Omega$ for asymptotic lower bounds.

## Big-Theta shortcuts

- Symmetry Rule:  If $f \in\Theta(g)$, then $g\in \Theta(f)$. 
- Reflexivity Rule:  $f \in \Theta(f)$. 
- Transitivity Rule:  If $f\in \Theta(g)$ and $g\in \Theta(h)$, then $f\in \Theta(h)$. 
- Sum Rule:  If $f\in O(g)$, then $f+g \in \Theta(g)$. 
- Constant Multiple Rule:  If $c>0$ is a constant, then $cf \in \Theta(f)$. 
- Product Rule:  If $f_1\in \Theta(g_1)$ and $f_2\in\Theta(g_2)$, then $f_1 \cdot f_2 \in \Theta(g_1 \cdot g_2)$.

## Little-oh and little-omega

- **$f(n)$ is $o(g(n))$** if $f(n)$ is $O(g(n))$ but not $\Theta(g(n))$.
    - Little-$o$ gives an upper bound that is not "tight".
    - Alternatively: $\displaystyle{\lim_{n\rightarrow \infty} \frac{f(n)}{g(n)} = 0}$
    
- **$f(n)$ is $\omega(g(n))$** if $f(n)$ is $\Omega(g(n))$ but not $\Theta(g(n))$.
    - Little-$o$ gives a lower bound that is not "tight".
    - Alternatively: $\displaystyle{\lim_{n\rightarrow \infty} \frac{f(n)}{g(n)} = \infty}$
    
## Join your table group

Join the *voice channel* for your group. (Video on if possible.)

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Timothy", "Josiah", "Nathan")
set.seed(1112021) 
n <- length(roster)
ngps <- 6
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

Complete the group exercise on the [Jamboard](). 

## Asymptotic notation practice

For each pair of expressions $(A,B)$, indicate whether $A$ is $O,o,\Omega,\omega,$ and/or $\Theta$ of $B$. Check all boxes that apply. Put a ? where you are not sure. Here $c>1$, $\epsilon>0$, and $k\geq 1$ represent constants.

| $A$            | $B$            | $O$ | $o$ | $\Omega$ | $\omega$ | $\Theta$  | 
|----------|----------|----------|----------|----------|----------|----------|
| $100n^2 + 1000n$ | $n^3$  |     |     |          |          |           |
| $\log_2 n$ | $\log_c n$  |     |     |          |          |           |
| $n^k$          | $c^n$          |     |     |          |          |           |
| $\sqrt{n}$     | $n^{\sin n}$   |     |     |          |          |           |
| $2^n$          | $2^{n/2}$      |     |     |          |          |           |
| $n^{\log_2 c}$ | $c^{\log_2 n}$ |     |     |          |          |           |
| $\log_2(n!)$   | $\log_2(n^n)$  |     |     |          |          |           |
| $(\log_2 n)^k$ | $n^\epsilon$   |     |     |          |          |           |

# Class Structure

## Daily Assignments on Canvas

- Due night before each class, 11:59pm.
- Questions posted in an RMarkdown file.
- Edit the RMarkdown file using RStudio, and write up solutions.
    - Use code blocks for pseudocode.
    - Use LaTeX for mathematics.
    - Markdown syntax works.
    - RMarkdown can also execute R code blocks.
- Knit to HTML, and upload `.html` file to Canvas.    
    
## Class Routine

- Goal is to spend the majority of the time working in groups on problems.
- Just-in-time mini-lectures.
- Some slides, but they won't be comprehensive.
- Can start by discussing assignment that was due the night before.

## Exams and Grading

| Task | % of total |
|----------|----------|
| Daily Assignments: | 44% |
| Exams: | 3 @ 14% each |
| Final Exam: | 14% |

## Attendance

- I expect every student to attend every class during the scheduled class period.  
- If you miss a significant number of classes, you will almost definitely do poorly in this class. 
- I consider it excessive to miss more than three classes during the course of the semester. 
- If you miss more than six classes without a valid excuse, I reserve the right to terminate you from the course with a grade of F.

## Academic integrity

- Dishonesty of any kind may result in loss of credit for the work involved and the filing of a report with the Provost’s Office.  
- Major or repeated infractions may result in dismissal from the course with a grade of F. 
- Be familiar with the [College’s plagiarism policy](https://www.westmont.edu/office-provost/academic-program/academic-integrity-policy).
- **Do not email, post online, or otherwise disseminate** any of the work that you do in this class.
- You may work with others on the assignments, but make sure that you type up your own answers yourself. 
- You are on your honor that the work you hand in represents your own understanding.

## Course Materials on GitHub

https://djhunter.github.io/algorithms/

- Assignments and slides posted here.
- Links to syllabus and textbook.
- Also source code for everything, if you want to look at the repository.
