---
title: "Divide and Conquer"
date: "January 15, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## Baguenaudier solver

## Recurrence Relation

## Closed-form formula

## Inductive verification

# Divide and Conquer

## Recall: Writing recursive functions

> Your **only** task is to simplify the original problem, or to solve it directly when simplification is either unnecessary or impossible; the *Recursion Fairy* will solve all the simpler subproblems for you, using Methods That Are None Of Your Business...

## Divide and Conquer

- When the problem domain *partitions* into smaller subproblems, the recursive paradigm is called **divide and conquer**.
    - Towers of Hanoi isn't really divide and conquer.
    - Binary search *is* divide and conquer. The search space partitions into two halves.

## Merge Sort

```
Merge(A[1..n], m):                        MergeSort(A[1..n]):
    i <- 1; j <- m + 1                        if n > 1
    for k <- 1 to n                               m <- floor(n/2)
        if j > n                                  MergeSort(A[1..m])
            B[k] <- A[i]; i <- i + 1              MergeSort(A[(m+1)]..m)
        else if i > m                             Merge(A[1..n],m)
            B[k] <- A[ j]; j <- j + 1
        else if A[i] < A[j]
            B[k] <- A[i]; i <- i + 1
        else
            B[k] <- A[j]; j <- j + 1
    for k <- 1 to n
        A[k] <- B[k]
```

## Merge Sort recurrence

```
Merge(A[1..n], m):   # O(n) comparisons   MergeSort(A[1..n]):  # C(n) comparisons
    i <- 1; j <- m + 1                        if n > 1
    for k <- 1 to n                               m <- floor(n/2)
        if j > n                                  MergeSort(A[1..m])      # C(n/2) comparisons (approx)
            B[k] <- A[i]; i <- i + 1              MergeSort(A[(m+1)]..m)  # C(n/2) comparisons (approx)
        else if i > m                             Merge(A[1..n],m)        # O(n) comparisons
            B[k] <- A[ j]; j <- j + 1
        else if A[i] < A[j]
            B[k] <- A[i]; i <- i + 1
        else
            B[k] <- A[j]; j <- j + 1
    for k <- 1 to n
        A[k] <- B[k]
```

$$
C(n) = \left\{\begin{array}{ll} 
               0 & \mbox{if } n=1 \\
               2C(n/2) + O(n) & \mbox{if } n>1
               \end{array}
              \right.
$$

## Merge Sort time

$$
C(n) = \left\{\begin{array}{ll} 
               0 & \mbox{if } n=1 \\
               2C(n/2) + O(n) & \mbox{if } n>1
               \end{array}
              \right.
$$

As a tree, we can model $C(n)$ as:

```
          O(n)
         /    \
     C(n/2)   C(n/2)
```

Expanding recursively, we get the following *recursion tree:*

```
                             n
                    /                 \
                n/2                      n/2
            /        \              /          \
          n/4        n/4           n/4          n/4
        /    \       /    \       /   \        /    \
      n/8    n/8   n/8   n/8    n/8   n/8    n/8    n/8
    ... and so on
```

## Merge Sort time

$$
C(n) = \left\{\begin{array}{ll} 
               0 & \mbox{if } n=1 \\
               2C(n/2) + O(n) & \mbox{if } n>1
               \end{array}
              \right.
$$

Total up each level in the recursion tree:

```
                                                            level total
                             n                                   n
                    /                 \
                n/2                      n/2                     n
            /        \              /          \ 
          n/4        n/4           n/4          n/4              n
        /    \       /    \       /   \        /    \
      n/8    n/8   n/8   n/8    n/8   n/8    n/8    n/8          n
    ... and so on
```

There are $\log n$ levels in this tree, so $C(n) \in O(n \log n)$.

(More to come.)

# Another Hanoi Puzzle

## Jamboard Questions

Consider the usual Towers of Hanoi puzzle (pp. 24ff) with $n$ disks and 3 pegs, numbered 0, 1, and 2. Now suppose you are forbidden to move any disk directly between peg 1 and peg 2; *every* move must involve peg 0.

1. Give a pseudocode block for a recursive algorithm to solve this puzzle.

2. Explain why (prove that) your algorithm never moves a disk between peg 1 and peg 2. (Use a *strong* induction hypothesis.)


## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Timothy", "Josiah", "Nathan")
set.seed(1152021) 
n <- length(roster)
ngps <- 6
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```
