---
title: "Divide and Conquer"
date: "January 15, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## More tips on describing algorithms (from pp. 11ff)

>- **Algorithms are not programs.** Please don't write code (e.g., C, Python, Java, Haskell, R, ...). 
>    - Use a combination of pseudocode and structured English. 
>        - Pseudocode uses the structure of formal programming languages to break algorithms into primitive steps.
>        - The primitive steps themselves can be written using math, English, or a mixture of the two, whatever is clearest.
>    - A good algorithm description is closer to what we should write in the comments of a real program than the code itself. Code is a poor medium for storytelling.
>    - On the other hand, conditionals, loops, function calls, and recursion should be stated explicitly.

## Who is the audience?

>- Imagine you are writing pseudocode for a competent but skeptical programmer who is *not as clever as you are*.
>    - This programmer should be able to quickly and correctly implement the algorithm in *their* favorite programming language.
>- **Goal:** include every detail necessary to fully specify the algorithm, prove its correctness, and analyze its running time.


## Baguenaudier solver

Pseudocode and English: It is clear what it's doing and *why it works*:

```
solve(n):

If n=1, 
    flip(1)
If n=2
    flip(2)
    flip(1)
else (when n>2)
    solve(n-2) on the rightmost n-2 bits
    flip(n), the nth bit                            << legal, b/c there are n-2 0's at the end >>
    solve(n-2) in reverse on the rightmost n-2 bits << makes Puzzle look like 0111...111 >>
    solve(n-1) on rightmost (n-1) bits
```

## Baguenaudier solver, but a little too code-y

Psuedocode only. It's correct, but it doesn't communicate as well:

```
B(int n){
        if(n == 0){
            print nothing
        }
        else if (n==1){
            print 1
        }
        else {
            B(n-2)
            print n
            B(n-2), but printed backwards 
            B(n-1)
        }
    }
```

## Recurrence Relation

The recurrence relation should match the algorithm:

```
solve(n):

If n=1, 
    flip(1)
If n=2
    flip(2)
    flip(1)
else (when n>2)
    solve(n-2) on the rightmost n-2 bits
    flip(n), the nth bit                            << legal, b/c there are n-2 0's at the end >>
    solve(n-2) in reverse on the rightmost n-2 bits << makes Puzzle look like 0111...111 >>
    solve(n-1) on rightmost (n-1) bits
```

\begin{equation}
  M(n)=\begin{cases}
    1 & \text{if $n=1$}.\\
    2 & \text{if $n=2$}.\\
    2M(n-2)+M(n-1)+1 & \text{if $n > 2$}.
  \end{cases}
\end{equation}

## Another Recurrence Relation

If the algorithm is different, the recurrence is different.

```
B(int n){
        if(n == 0){
            print nothing
        }
        else if (n==1){
            print 1
        }
        else {
            B(n-2)
            print n
            B(n-2), but printed backwards 
            B(n-1)
        }
    }
```

\begin{equation}
  M(n)=\begin{cases}
    0 & \text{if $n=0$}.\\
    1 & \text{if $n=1$}.\\
    2M(n-2)+M(n-1)+1 & \text{if $n > 2$}.
  \end{cases}
\end{equation}

## Closed-form formula from OEIS

At least two options:

$$
2^{n+1}/3 - 1/2 - (-1)^n/6
$$
or 

$$
M\left(n\right) = \left\lceil2\cdot\frac{2^n-1}{3}\right\rceil
$$
The former is easier; the latter requires cases in the induction argument.

## Inductive verification

Base case: $M(1)=1=\frac{2^{2}}{3}-\frac{1}{2}-\frac{(-1)^{1}}{6}$ and $M(2)=2=\frac{2^{3}}{3}-\frac{1}{2}-\frac{(-1)^{2}}{6}$.

Suppose as inductive hypothesis that $M(k)=\frac{2^{k+1}}{3}-1/2-\frac{(-1)^{k}}{6}$ for all $k<n$.

(When in doubt, always use *strong* induction.)

## Inductive step

\begin{align}
M(n)&=2M(n-2)+M(n-1)+1\\
&=2\left(\frac{2^{n-1}}{3}-\frac{1}{2}-\frac{(-1)^{n-2}}{6} \right) + \left(\frac{2^{n}}{3}-\frac{1}{2}-\frac{(-1)^{n-1}}{6}\right)+1\\
&=\frac{2^{n}}{3}-1-\frac{2(-1)^{n-2}}{6}+\frac{2^{n}}{3}-\frac{1}{2}-\frac{(-1)^{n-1}}{6}+1\\
&=\frac{2^{n}}{3}+\frac{2^{n}}{3}-1+1-\frac{1}{2}-\frac{2(-1)^{n-2}}{6}-\frac{(-1)^{n-1}}{6}\\
&=\frac{2^{n+1}}{3}-\frac{1}{2}+\frac{2(-1)^{n-1}}{6}-\frac{(-1)^{n-1}}{6}\\
&=\frac{2^{n+1}}{3}-\frac{1}{2}+\frac{(-1)^{n-1}}{6}\\
&=\frac{2^{n+1}}{3}-\frac{1}{2}-\frac{(-1)^{n}}{6}
\end{align}

# Trust the Recursion Fairy

## Recall: Writing recursive functions

> Your **only** task is to simplify the original problem, or to solve it directly when simplification is either unnecessary or impossible; the *Recursion Fairy* will solve all the simpler subproblems for you, using Methods That Are None Of Your Business...

## Jamboard Questions

Consider the usual Towers of Hanoi puzzle (pp. 24ff) with $n$ disks and 3 pegs, numbered 0, 1, and 2. Now suppose you are forbidden to move any disk directly between peg 1 and peg 2; *every* move must involve peg 0.

1. Describe a recursive algorithm to solve this puzzle.

2. Explain why (prove that) your algorithm never moves a disk between peg 1 and peg 2. (Use a *strong* induction hypothesis.)


## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Timothy", "Josiah", "Nathan")
set.seed(1152021) 
n <- length(roster)
ngps <- 6
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

# Divide and Conquer

## Divide and Conquer paradigm

- When the problem domain *partitions* into smaller subproblems, the recursive paradigm is called **divide and conquer**.
    - Towers of Hanoi isn't really divide and conquer.
    - Binary search *is* divide and conquer. The search space partitions into two halves.

## Merge Sort

```
Merge(A[1..n], m):                        MergeSort(A[1..n]):
    i <- 1; j <- m + 1                        if n > 1
    for k <- 1 to n                               m <- floor(n/2)
        if j > n                                  MergeSort(A[1..m])
            B[k] <- A[i]; i <- i + 1              MergeSort(A[(m+1)]..m)
        else if i > m                             Merge(A[1..n],m)
            B[k] <- A[ j]; j <- j + 1
        else if A[i] < A[j]
            B[k] <- A[i]; i <- i + 1
        else
            B[k] <- A[j]; j <- j + 1
    for k <- 1 to n
        A[k] <- B[k]
```

## Merge Sort recurrence

```
Merge(A[1..n], m):   # O(n) comparisons   MergeSort(A[1..n]):  # C(n) comparisons
    i <- 1; j <- m + 1                        if n > 1
    for k <- 1 to n                               m <- floor(n/2)
        if j > n                                  MergeSort(A[1..m])      # C(n/2) comparisons (approx)
            B[k] <- A[i]; i <- i + 1              MergeSort(A[(m+1)]..m)  # C(n/2) comparisons (approx)
        else if i > m                             Merge(A[1..n],m)        # O(n) comparisons
            B[k] <- A[ j]; j <- j + 1
        else if A[i] < A[j]
            B[k] <- A[i]; i <- i + 1
        else
            B[k] <- A[j]; j <- j + 1
    for k <- 1 to n
        A[k] <- B[k]
```

$$
C(n) = \left\{\begin{array}{ll} 
               0 & \mbox{if } n=1 \\
               2C(n/2) + O(n) & \mbox{if } n>1
               \end{array}
              \right.
$$

## Merge Sort time

$$
C(n) = \left\{\begin{array}{ll} 
               0 & \mbox{if } n=1 \\
               2C(n/2) + O(n) & \mbox{if } n>1
               \end{array}
              \right.
$$

As a tree, we can model $C(n)$ as:

```
          O(n)
         /    \
     C(n/2)   C(n/2)
```

Expanding recursively, we get the following *recursion tree:*

```
                             n
                    /                 \
                n/2                      n/2
            /        \              /          \
          n/4        n/4           n/4          n/4
        /    \       /    \       /   \        /    \
      n/8    n/8   n/8   n/8    n/8   n/8    n/8    n/8
    ... and so on
```

## Merge Sort time

$$
C(n) = \left\{\begin{array}{ll} 
               0 & \mbox{if } n=1 \\
               2C(n/2) + O(n) & \mbox{if } n>1
               \end{array}
              \right.
$$

Total up each level in the recursion tree:

```
                                                            level total
                             n                                   n
                    /                 \
                n/2                      n/2                     n
            /        \              /          \ 
          n/4        n/4           n/4          n/4              n
        /    \       /    \       /   \        /    \
      n/8    n/8   n/8   n/8    n/8   n/8    n/8    n/8          n
    ... and so on
```

There are $\log n$ levels in this tree, so $C(n) \in O(n \log n)$.

(More to come.)
