---
title: "Recursion Trees"
date: "January 22, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## Balance between informal and formal

- Include enough details to be correct.
- Try not to include details that obfuscate.
- Strive for an argument that *convinces* someone that the algorithm really works.
    - A description of what the algorithm does is not a proof that the algorithm is correct.

## Cavarretta Sort Correctness

```
CavarrettaSort(A[0 .. n - 1]):
    if n = 2 and A[0] > A[1]
        swap A[0] and A[1]
    else if n > 2
         m = ceiling(2n/3)
         CavarrettaSort(A[0 .. (m - 1)])
         CavarrettaSort(A[(n - m) .. (n - 1)])
         CavarrettaSort(A[0 .. (m - 1)])
```

## Clear proof, with a couple details glossed over

Base case: If n = 2, the two elements of the list are swapped if out of order, resulting a correctly sorted list.

Suppose as inductive hypothesis that `CavarrettaSort(A[0 .. k-1])` correctly sorts its input array for all $k<n$.

Then `CavarrettaSort(A[0...n-1])` takes the following steps: 1. Correctly sorts the first 2/3 of the array, according to the inductive hypothesis 2. Correctly sorts the last 2/3 of the array, according to the inductive hypothesis 3. Correctly sorts the first 2/3 of the array again, according to the inductive hypothesis.

After step 2, the last 1/3 of the array will contain the highest 1/3 of numbers, because any of these numbers that may have been in the first 1/3 were moved into the middle 1/3 in step 1, and the last 1/3 in step 2. After step 3, the first 2/3 of the array will be sorted correctly as well, because all numbers in the first 2/3 at this point are less than all numbers in the last 1/3, resulting in a complete solution.

## Alternative inductive step (more details)

First, since $m$ is the *ceiling* of $2n/3$, $m-n$ is never larger than $2m-n$. 

After the first `CavarrettaSort(A[0 .. (m - 1)])` call, all of the $n-m$ elements of `A[(2m-n)..(m-1)]` (the second "1/3") are greater than all of the $2m-n$ elements of `A[0..(2m-n-1)]` (the first "1/3"), by inductive hypothesis. Therefore, *none of the largest $n-m$ elements of the entire array can be in `A[0..(n-m-1)]`,* and so the largest $n-m$ elements must be in `A[(n-m)..(n-1)]` (the last "2/3"). 

After the call `CavarrettaSort(A[(n - m) .. (n - 1)])`, these largest $n-m$ elements must be in order in `A[m..(n-1)]` (the last "1/3"), by inductive hypothesis. The final recursive call puts the smallest $m$ elements of the array in order in `A[0..(m-1)]`, by inductive hypothesis. Therefore, the entire array is correctly sorted.

## Recurrence

```
CavarrettaSort(A[0 .. n - 1]):
    if n = 2 and A[0] > A[1]
        swap A[0] and A[1]
    else if n > 2
         m = ceiling(2n/3)
         CavarrettaSort(A[0 .. (m - 1)])
         CavarrettaSort(A[(n - m) .. (n - 1)])
         CavarrettaSort(A[0 .. (m - 1)])
```

$$
C(n) = \left\{\begin{array}{lr}  1, 
& \text{for } n=2,\\  
3C\left(\left\lceil\frac{2}{3}n\right\rceil\right), 
& \text{for } n>2.\\  \end{array}\right.
$$

## Solving the recurrence

Ignoring the ceiling,

\begin{align}
C(n) &= 3C\left(\frac{2}{3}n\right) = 3^2C\left(\left(\frac{2}{3}\right)^2n\right) 
= 3^3C\left(\left(\frac{2}{3}\right)^2n\right) \\ 
& = \cdots = 3^LC\left(\left(\frac{2}{3}\right)^Ln\right) = 3^L C(2) = 3^L \cdot 1 = 3^L
\end{align}

Where $\left(\frac{2}{3}\right)^Ln = 2$. Solving for $n$ gives $L = \log_{1.5}(n/2)$. 

Therefore, $C(n) = 3^{\log_{1.5}(n/2)} = (n/2)^{\log_{1.5} 3} \in O(n^{\log_{1.5} 3}) = O(n^{2.7095\ldots})$

# Recursion Trees

## Recall: Merge (Quick?) Sort Recurrence

$$
C(n) = \left\{\begin{array}{ll} 
               0 & \mbox{if } n=1 \\
               2C(n/2) + O(n) & \mbox{if } n>1
               \end{array}
              \right.
$$

Total up each level in the recursion tree:

```
                                                            level total
                             n                                   n
                    /                 \
                n/2                      n/2                     n
            /        \              /          \ 
          n/4        n/4           n/4          n/4              n
        /    \       /    \       /   \        /    \
      n/8    n/8   n/8   n/8    n/8   n/8    n/8    n/8          n
    ... and so on
```

There are $\log n$ levels in this tree, so $C(n) \in O(n \log n)$.

## General divide and conquer relations

- Each step costs $O(f(n))$, plus
- the cost of $r$ subproblems, each of size $n/c$.
- We want an asymptotic (big-O) solution

$$
T(n) = \left\{\begin{array}{ll} 
               \mbox{doesn't matter} & \mbox{if } n \leq n_0 \\
               rT(n/c) + O(f(n)) & \mbox{if } n > n_0
               \end{array}
              \right.
$$
Recursion tree:

- $r$-way branches ($r$-ary tree)
- Level of leaves: $L = \log_c n$
- Number of leaves: $r^L = r^{\log_c n} = n^{\log_c r}$

See Figure 1.9.

## Three cases you can solve

Recursive case of divide and conquer recurrence:

$$
T(n) = rT(n/c) + O(f(n))
$$

1. Root dominates: $T(n) = O(f(n))$
    - happens when row sums *decrease* geometrically.
2. Internal nodes dominate: $T(n) = O(f(n)\log n)$
    - happens when row sums are *constant*.
3. Leaves dominate: $T(n) = O(n^{\log_c r})$
    - happens when row sums *increase* geometrically.
    
[CLRS] calls this the **Master Theorem** for solving d&c recurrences.

## Examples

>- $C(n) = 2C(n/2) + O(n)$, (Merge Sort)
>    - row sums all equal $n$. Case 2: $C(n) = O(n \log n)$
>- $D(n) = 4D(n/2) + O(n^3)$
>    - row sums decrease geometrically. Case 1: $D(n) = O(n^3)$
    
## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(1222021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```


## Recursion Tree Practice

1. Draw a recursion tree for the recurrence $F(n) = 3F(n/4) + O(\sqrt{n})$.

2. Write down the first four terms in the sequence of row sums.

3. Apply the Master Theorem to solve the recurrence.

## Cavarretta recurrence, revisited.

4. Consider adding a constant term $K = O(1)$  to the Cavarretta recurrence:

$$
C(n) = \left\{\begin{array}{lr}  1, 
& \text{for } n=2,\\  
3C\left(\left\lceil\frac{2}{3}n\right\rceil\right) + K, 
& \text{for } n>2.\\  \end{array}\right.
$$

Use a recursion tree (with root $K$) to solve this recurrence. What effect did adding the constant term have?
