---
title: "More Backtracking Algorithms"
date: "January 29, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## General comments

## ChurroCut Time

## ChurroCut Timing

# More Backtracking Examples

## General Idea

- Problem has several options, and *recursive substructure.*
- Try one of the options, and solve the subproblem recursively.
    - If that doesn't work, try another option.
    
Backtracking algorithms tend to be slow.

## Decision Problems

- A *decision problem* is a problem whose answer is either TRUE or FALSE.
- Lots of basic questions about algorithms involve decision problems.
- Often an algorithm that solves a decision problem can be modified slightly to do something that seems harder.

Example: 

- Decision problem: Does $p(x) = Ax^2 + Bx +C$ have any roots?
- Slightly harder problem: Find all the roots of $p(x) = Ax^2 + Bx +C$.

Meta-Example:

- Decision problem: Does $\mathcal{X}$ have a solution?
- Slightly harder problem: Give all the solutions of $\mathcal{X}$.

# Text Segmentation

## Example: Text Segmentation

- We have a boolean function `IsWord(A[1..i])` that tells us if a string (i.e., array) is a word in our language.
- We have a string of words without spaces.
- Can the string be *segmented* into words? (Is it "splittable"?)

```
BOTHEARTHANDSATURNSPIN

BOT HEART HANDS AT URNS PIN

BOTH EARTH AND SATURN SPIN 
```

## Recursive substructure

If a string starts with a word, and the rest of the string is splittable into words, then whole string is splittable.

- See if the first letter is a word.
    - Recursively check the rest.
- See if the first two letters make a word.
    - Recursively check the rest.
- See if the first three letters make a word.
    - Recursively check the rest.
- ... and so on.

## Text Segmentation Algorithm

```
Splittable(A[1..n]):
    if n = 0
        return TRUE
    for i <- 1 to n
        if IsWord(A[1..i])
            if Splittable(A[(i+1)..n])
                return TRUE
    return FALSE
```

Time? (Same recurrence as CutChurro: $O(2^n)$).

## Global Variables

- Is usually *not OK* to use global variables when programming.
- It is actually **OK** to use global variables when describing algorithms.
    - Often global variables make our descriptions cleaner.
    - We assume a "competent programmer who is not quite as clever as we are" could read our algorithm description and implement it without global variables.
    
## Text Segmentation Algorithm: Index only version

```
A[1..n] is a string to test to see if it is splittable.
IsWord(i,j) returns true iff A[i..j] is a word.

Splittable(i):    << Is A[i..n] splittable? >>
    if i > n
        return TRUE
    for j <- i to n
        if IsWord(i, j)
            if Splittable(j+1)
                return TRUE
    return FALSE
    
    
Call this function as: Splittable(n)
```

# Subset Sum

## Example: Subset Sum

- We have a *set* $X$ of positive integers (so no duplicates).
- Given a *target* value $T$, is there some *subset* of $X$ that adds up to $T$?

Recursive Substructure:

- For any $x\in X$, the answer is TRUE for target $T$ if
    - The answer is TRUE for $X \setminus \{x\}$ and target $T$, or
    - The answer is TRUE for $X \setminus \{x\}$ and target $T-x$.
- In other words, the cases are:
    - *without:* There is a subset not containing $x$ that works.
    - *with:* There is a subset containing $x$ that works.
- Base Cases:
    - If $T = 0$, the answer is TRUE. ($\emptyset$ works!)
    - If $T < 0$, the answer is FALSE. (everything's positive)
    - If $X = \emptyset$, the answer if FALSE (unless $T=0$, of course.)

## Subset Sum Algorithm: Set version

Form the subproblems by removing *some* element of $X$.

```
SubsetSum(X, T):   <<  Does any subset of X sum to T?  >>
    if T = 0
        return TRUE
    else if T < 0 or X = ∅
        return FALSE
    else
        x <- any element of X
        with <- SubsetSum(X \ {x}, T − x)
        wout <- SubsetSum(X \ {x}, T)
        return (with OR wout)
```

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(1292021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

## Subset Sum: Array index version

1. Complete the array index version of `SubsetSum`. Form the subproblems by removing the *last* element of $X$.

```
X[1..n] is a global array of positive integers

SubsetSum(i, T) :  <<  Does any subset of X[1..i] sum to T?  >>
{
    TODO
}


Call this function as: SubsetSum(n, T)
```

2. Come up with a recurrence for the number of `SubsetSum` calls made by `SubsetSum(n)`. Do you recognize it? What's the closed-form solution?

## Slightly Harder Problem

3. Can you modify the algorithm from #1 so that it returns the *number* of subsets of $X$ that add up to $T$, instead of just TRUE or FALSE?

