---
title: "Dynamic Programming: Introduction"
date: "February 1, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## Counting Subsets

```{javascript, eval=FALSE}
SubsetSum(i,T):    //  Does any subset of X[1..i] sum to T?  
  if T = 0
    return 1
  else if T < 0 or i = 0
    return 0
  else
    with <- SubsetSum(i-1, T − X[i])
    without <- SubsetSum(i-1, T) 
    return (with + without)
```

- Base case: If $T=0$, there is one subset, $\emptyset$, summing to $T$.
- Inductive hypothesis: `SubsetSum(k,T)` returns the number of subsets of `X[1..k]` summing to $T$, for $k<i$.
- Inductive step: There are two *disjoint* ways the target can be reached: using `X[i]` and not using `X[i]`, so the total number of ways is the sum of these two.

## Recall: Text Segmentation 

Break a string into words: `BOTHEARTHANDSATURNSPIN`

```{javascript, eval=FALSE}
Splittable(A[1..n]):
    if n = 0
        return TRUE
    for i <- 1 to n
        if IsWord(A[1..i])
            if Splittable(A[(i+1)..n])
                return TRUE
    return FALSE
```

## Text Segmentation (array index version)

```{javascript, eval=FALSE}
// A[1..n] is a string to test to see if it is splittable. (n is constant)
// IsWord(i,j) returns true iff A[i..j] is a word.

Splittable(i):    // Is A[i..n] splittable? 
    if i > n
        return TRUE
    for j <- i to n
        if IsWord(i, j)
            if Splittable(j+1)
                return TRUE
    return FALSE
    
    
// Call this function as: Splittable(1)
```

## Counting word segmentations

```{javascript, eval=FALSE}
// A[1..n] is a string (n is constant)
// IsWord(i,j) returns true iff A[i..j] is a word.

PartitionCount(i):   // how many ways can A[i..n] split into words?
    if i > n
        return 1 
    total <- 0
    for j <- i to n
        if IsWord(i, j)
            total <- total + PartitionCount(j+1)
    return total
```

- Base Case: There are no letters to consider, so we have one vacuous split.
- Inductive hypothesis: `PartitionCount(k)` returns the number of splits of `A[k..n]` for $k > i$.
- Inductive Step: Check all the possible prefixes of `A[i..n]` for a word. For each that is found, increment `total` by the number of splits of the remainder, which is correctly determined, by inductive hypothesis.

## Recursion Tree?

Consider the recursion tree for `TOPARTISTOIL`.

# Dynamic Programming

## Why is backtracking so slow?

- Problem has several options, and *recursive substructure.*
- Many of the subproblems are the same: *overlapping subproblems*.
    - A brute-force recursive algorithm will repeatedly solve the same problem.
    
**Solution:** Save the subproblem calculations (usually in an array), and refer to the array instead of making a recursive call.
    
## Example: Fibonacci Numbers

$$
F(n) = \left\{\begin{array}{ll}
1 & \mbox{if } n = 0 \mbox{ or } n = 1 \\
F(n-1) + F(n-1) & \mbox{if } n>1
\end{array} \right.
$$
Brute force recursion (exponential time):

```
RFib(n):
    if n = 0
        return 0
    else if n = 1
        return 1
    else
    return RFib(n − 1) + RFib(n − 2)
```

Top-down evaluation wastes a lot of effort:

```
RFib(7) =                   RFib(6)                               + RFib(5)
           =       RFib(5)          + RFib(4)            + RFib(4)          + RFib(3)
           = RFib(4) + RFib(3) + RFib(3) + RFib(2) + RFib(3) + RFib(2) + RFib(2) + RFib(1)
           = etc...
```

## Better: Memoized Recursion

**Memoization:**  Avoid repeated recursive calls by storing every result in an array.
Instead of making a recursive call, just use the array value (if defined).

```
MemFibo(n):
    if n = 0
        return 0
    else if n = 1
        return 1
    else
        if F[n] is undefined
            F[n] <- MemFibo(n − 1) + MemFibo(n − 2)
        return F[n]
```

- The recursion "tree" is trimmed: constant "width"
- The algorithm is actually iterative.
    - $F[n]$ gets populated from the bottom up: $F[0], F[1], F[2], \ldots$

## Even Better: Iterative Solution

Notice: 

- $F[n]$ gets populated from the bottom up: $F[0], F[1], F[2], \ldots$
- So loop from bottom up, and populate $F$ as you go.
    - Populate $F$ recursively, but in a loop.
    
```
IterFibo(n):
    F[0] <- 0
    F[1] <- 1
    for i <- 2 to n
        F[i] <- F[i − 1] + F[i − 2]
    return F[n]
```

- Time: $\Theta(n)$
- Space: $\Theta(n)$ 
    - But we could make it $\Theta(1)$ by just keeping last two $F$'s as we go.
    
# Writing Dynamic Programming Algorithms

## Dynamic Programming Steps

1. *Specify* the problem as a recursive function, and *solve it* recursively (with an algorithm or formula).

2. *Build the solutions* from the bottom up:
    - What are the *subproblems*?
    - What *data structure* do I store the solutions in? (Usually an array.)
    - Which subproblem does each subproblem depend on?
        - Determine an *evaluation order.*
    - Space and time?
    - Write the algorithm. Usually it will just be a couple nested loops, with the recursive calls replaced by array look-ups.

## Counting word segmentations

```{javascript, eval=FALSE}
// A[1..n] is a string (n is constant)
// IsWord(i,j) returns true iff A[i..j] is a word.

PartitionCount(i):   // how many ways can A[i..n] split into words?
    if i > n
        return 1 
    total <- 0
    for j <- i to n
        if IsWord(i, j)
            total <- total + PartitionCount(j+1)
    return total
    
    
// Call this function as PartitionCount(1)
```

Time: $O(2^n)$

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(212021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

## Using dynamic programming

1. *Identify the subproblems.* In order for `PartitionCount(i)` to return a value, which subproblems need to return values? (i.e., what does the Recursion Fairy need to take care of?)

2. *Find an evaluation order.* Let `PC` be an array of numbers for filling in the solutions, so our job is to put the return value of `PartitionCount(i)` in `PC[i]`, for all `i`. What is the first value of `PC` that you can fill in? Based on that first value, what's the second value you can fill in? Third?

3. Write an iterative *algorithm* to fill in the elements of `PC`.

4. What are the *space and time* of your algorithm?

