---
title: "Dynamic Programming: Tables"
date: "February 2, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## Pseudocode is not code

```{javascript, eval=FALSE}
// Let price[1..N] be a constant array of prices.
// So price[i] is the cost of a churro of length i.

CutChurro(n):
    if n = 0
        return 0
    maxTotPrice <- -1
    for i <- 1 to n
        totPriceTry <- price[i] + CutChurro(n-i)
        if totPriceTry > maxTotPrice
            maxTotPrice <- totPriceTry 
    return maxTotPrice
```

- The goal is to *communicate* your algorithm in a way that makes its correctness transparent.
- It's OK to have things a little less "buttoned-up" if it makes it clearer what the algorithm is doing.

## Step 1: understand the recursive structure

```{javascript, eval=FALSE}
// Let price[1..N] be a constant array of prices.
// So price[i] is the cost of a churro of length i.

CutChurro(n):
    if n = 0
        return 0
    maxTotPrice <- -1
    for i <- 1 to n
        totPriceTry <- price[i] + CutChurro(n-i)
        if totPriceTry > maxTotPrice
            maxTotPrice <- totPriceTry 
    return maxTotPrice
```

- `CutChurro(n)` depends on the Recursion Fairy taking care of `CutChurro(n-1)`, `CutChurro(n-2)`, ..., `CutChurro(0)`.
- Data structure: One-dimensional array `CC[0..n]`
- Evaluation order: `CC[0]` is the first thing to fill in, then `CC[1]`, `CC[2]`, ... etc, in that order.
    
## Fast Churro Cutting

```{javascript, eval=FALSE}
FastCutChurro(n):
    CC[0] <- 0
    for i <- 1 to n
        maxTotPrice <- 0
        for j <- 1 to i
            totPriceTry <- price[j] + CC[i-j]
            if maxTotPrice < totPriceTry
                maxTotPrice <- totPriceTry
        CC[i] <- maxTotPrice
    return CC[n]
```
    
- The recurrence in this algorithm is transparently the same as the recursive call in the recursive algorithm.

## Fast Churro Cutting: Space and Time

```{javascript, eval=FALSE}
FastCutChurro(n):
    CC[0] <- 0
    for i <- 1 to n
        maxTotPrice <- 0
        for j <- 1 to i
            totPriceTry <- price[j] + CC[i-j]
            if maxTotPrice < totPriceTry
                maxTotPrice <- totPriceTry
        CC[i] <- maxTotPrice
    return CC[n]
```
    
- Old (recursive) version: $O(2^n)$ (exponential time)
- Dynamic programming (iterative) time: $O(n^2)$ (nested for-loops)
    - Space: $O(n)$, because of the one-dimensional array
- Since $n^2 \in o(2^n)$, this time is *subexponential*.
    - Exponential functions dominate polynomials.
    
## Fast Churro Cutting: Exponentially better time

```{javascript, eval=FALSE}
FastCutChurro(n):
    CC[0] <- 0
    for i <- 1 to n
        maxTotPrice <- 0
        for j <- 1 to i
            totPriceTry <- price[j] + CC[i-j]
            if maxTotPrice < totPriceTry
                maxTotPrice <- totPriceTry
        CC[i] <- maxTotPrice
    return CC[n]
```
    
- Actual timings show that going from $O(2^n)$ to $O(n^2)$ is very noticable, even for small examples:
    
```
> system.time(CutChurro(20))
   user  system elapsed 
  0.736   0.002   0.738 
> system.time(FastCutChurro(20))
   user  system elapsed 
      0       0       0 
```

## Coding Optimizations

- **Note:** Better *code* is not necessarily better *pseudocode*.
- These examples are cleaner, but the connection between the recursive algorithm is less transparent.

```{javascript, eval=FALSE}
FastCutChurro(n):
    CC[0] = 0
    for (i in 1:n)
        CC[i] = -1
        for (j in 1:i)
            p = price[j] + CC[i-j]
            if p > CC[i]: CC[i] = p
    return CC[n]
    

FastCutChurro(n):
    Initialize CC[1..n] to price[1..n]
    for i in 2..n:
        for j in 1..floor(i/2):
            if CC[j] + CC[i-j] > CC[i]:
                CC[i] <- CC[j] + CC[i-j]
    return CC[n]
```

# Dynamic Programming

## Dynamic Programming Problems

- Problem has several options, and *recursive substructure.*
    - String splitting: find a word, then split what's left.
    - Churro cutting: cut off a piece, then price what's left.
- Many of the subproblems are the same: *overlapping subproblems*.
    - The same substrings recur.
    - Several churro pieces have same length.

Dynamic programming keeps track of the subproblem solutions, and iteratively builds an array/table of solutions.

## Dynamic Programming Steps

1. *Specify* the problem as a recursive function, and *solve it* recursively (with an algorithm or formula).

2. *Build the solutions* from the bottom up:
    - What are the *subproblems*?
    - What *data structure* do I store the solutions in? (Usually an array.)
    - Which subproblem does each subproblem depend on?
        - Determine an *evaluation order.*
    - Space and time?
    - Write the algorithm. Usually it will just be a couple nested loops, with the recursive calls replaced by array look-ups.

# Edit Distance

## Minimum edit distance between two strings

How many operations does it take to transform one string (e.g., DNA sequence) to another? The operations are:

- Mutation: Change one letter to another value.
- Deletion: Delete one letter.
- Insertion: Insert a letter.

```
SATURDAY -mutate-> SUTURDAY -mutate-> SUNURDAY -delete-> SUNRDAY -delete-> SUNDAY
SATURDAY -delete-> STURDAY -delete-> SURDAY -mutate-> SUNDAY
KITTEN -mutate-> SITTEN -mutate-> SITTIN -insert-> SITTING
```

Alignment diagrams: |-bars represent 1 unit of cost.

```
SATURDAY    SATURDAY    KITTEN     ALGOR I THM
 ||||        || |       |   | |      || | | ||
SUN  DAY    S  UNDAY    SITTING    AL TRUISTIC
```

## Recursive Substructure

```
SATURDAY    SATURDAY    KITTEN     ALGOR I THM
 ||||        || |       |   | |      || | | ||
SUN  DAY    S  UNDAY    SITTING    AL TRUISTIC
```

The rightmost element in an alignment diagram can be:

- A match (cost 0) $\substack{x \\ | \\ x}$
- A mutation (cost 1) $\substack{x \\ | \\ y}$
- An insertion (cost 1) $\substack{- \\ | \\ x}$
- A deletion (cost 1) $\substack{x \\ | \\ -}$

Try all four, then have the Recursion Fairy optimize the rest.

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(232021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

## Use the Recursion Fairy

Suppose `A[1..m]` and `B[1..n]` are two strings. Suppose also that you have a function `EditDistance(i,j)` that returns the minimum edit distance from `A[1..i]` to `B[1..j]`, as long as $i<m$ or $j<n$.

1. Suppose `A[m] = B[n]` (the last characters of the strings match). If the rightmost element of the alignment diagram is a match $\substack{x \\ | \\ x}$, what is the minimum edit distance from `A[1..m]` to `B[1..n]`? (Use the `EditDistance` function.)

2. Suppose `A[m] != B[n]` (the last characters of the strings don't match). If the rightmost element of the alignment diagram is a mutation $\substack{x \\ | \\ y}$, what is the minimum edit distance?

## Insertions and Deletions

3. If the rightmost element of the alignment diagram is an insertion $\substack{- \\ | \\ x}$, what is the minimum edit distance?

4. If the rightmost element of the alignment diagram is a deletion $\substack{x \\ | \\ -}$, what is the minimum edit distance?

# Using Dynamic Programming

## Data Structure and recurrence

Let `ED[1..m, 1..n]` be an array in which we will store the minimum edit distance from `A[1..i]` to `B[1..j]`. 

`ED[i,j]` will be the minimum of the following:

- `ED[i-1, j-1]` (if `A[i] = B[j]`) 
    - match
- `ED[i-1, j-1] + 1` (if `A[i] != B[j]`)
    - mutation
- `ED[i, j-1] + 1` 
    - insertion
- `ED[i-1, j] + 1`
    - deletion

## Evaluation Order

`ED[i,j]` is the minimum of the following:

- `ED[i-1, j-1]` (if `A[i] = B[j]`) 
- `ED[i-1, j-1] + 1` (if `A[i] != B[j]`)
- `ED[i, j-1] + 1` 
- `ED[i-1, j] + 1`

So the evaluation order must obey the arrows:

$$
\begin{array}{cccccc}
\cdots & \mathtt{ED[i-2, j-2]} & \rightarrow & \mathtt{ED[i-2, j-1]} & \rightarrow &\mathtt{ED[i-2, j]} \\
 & \downarrow & \searrow & \downarrow & \searrow & \downarrow \\
\cdots & \mathtt{ED[i-1, j-2]} & \rightarrow & \mathtt{ED[i-1, j-1]} & \rightarrow &\mathtt{ED[i-1, j]} \\
 & \downarrow & \searrow & \downarrow & \searrow & \downarrow \\
\cdots & \mathtt{ED[i, j-2]} & \rightarrow & \mathtt{ED[i, j-1]} & \rightarrow & \mathtt{ED[i, j]} \\
\end{array}
$$

## Row major or column major

$$
\begin{array}{cccccc}
\cdots & \mathtt{ED[i-2, j-2]} & \rightarrow & \mathtt{ED[i-2, j-1]} & \rightarrow &\mathtt{ED[i-2, j]} \\
 & \downarrow & \searrow & \downarrow & \searrow & \downarrow \\
\cdots & \mathtt{ED[i-1, j-2]} & \rightarrow & \mathtt{ED[i-1, j-1]} & \rightarrow &\mathtt{ED[i-1, j]} \\
 & \downarrow & \searrow & \downarrow & \searrow & \downarrow \\
\cdots & \mathtt{ED[i, j-2]} & \rightarrow & \mathtt{ED[i, j-1]} & \rightarrow & \mathtt{ED[i, j]} \\
\end{array}
$$

- Either *row major* (row by row) or *column major* (column by column) order will work.
- For example, here is row major as a nested for-loop:

```
for i from 1 to m
    for j from 1 to n
         TODO: populate E[i,j]
```

## Space and Time

- *Data structure:* `ED[1..m, 1..n]`, so **space** is $O(mn)$.
- *Recursive structure:* `ED[i,j]` is the minimum of the following:
    - `ED[i-1, j-1]` (if `A[i] = B[j]`) 
    - `ED[i-1, j-1] + 1` (if `A[i] != B[j]`)
    - `ED[i, j-1] + 1` 
    - `ED[i-1, j] + 1`
        - This minimum can be taken in constant time ($O(1)$).
- *Evaluation order*:

```
for i from 1 to m
    for j from 1 to n
         TODO: populate E[i,j]
```

- The `populate` step takes constant time, so **time** is $mn \cdot O(1) = O(mn)$

## Algorithm

Finally, we can write the algorithm. (Note we add a row and a column of 0's as a base case.)

```{python, eval=FALSE}
EditDistance(A[1 .. m], B[1 .. n]):
    Initialize ED[0, *] and ED[*, 0] to *, for *=0,1,2,... 
    for i <- 1 to m
        for j <- 1 to n
            insert <- ED[i, j − 1] + 1
            delete <- ED[i − 1, j] + 1
            if A[i] = B[j]
                mutate <- ED[i − 1, j − 1]
            else
                mutate <- ED[i − 1, j − 1] + 1
            ED[i, j] <- min {insert, delete, mutate}
    return ED[m, n]
```

Check out the memoization table.

## Exam #1: February 10 (Wed.)

- Are you ready?
    - You should be able to do [tonight's assignment](../assignments/10changemaking.html) *by yourself*, without getting help from others.
        - It's OK to get help from others, but before you do you should try to solve the problem *by yourself*.
    - If you can't do this assignment *by yourself*, that's a sign that you are relying too much on others to do your work for you.
- What if you're not ready?
    - Get a notebook and a pencil and redo [all the assignments](../index.html#Written_Assignments) *by yourself* without looking at the solutions.
    - Try some other Exercises in the book. The goal is to *exercise* the techniques you have learned, not necessarily produce a perfect solution.
    
