---
title: "Dynamic Programming: Subset Sum"
date: "February 5, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## Working Together

- It's fine to work together, as long as the work you turn in represents work that you understand and participated in.
- Consider not always working with the same group.
- *Invite* other to join you; Feel free to use the `#CS-120` text channel to announce when you'll be working.

For example,

> Hey everyone! I'm planning to be in VC tonight around 7 to work on the assignment. Anyone want to join me?

Feel free to use the voice channels in the Winter Hall Learning Lounge server.

## Describing Algorithms

- Go back and read Section 0.5 again. It might make more sense now that we have a little more context.
- Remember: the goal is to *communicate*.

## Change Making Examples

```
dollarValues = [1, 4, 7, 13, 28, 52, 91, 365]

MinDreamDollars(k):
  # base case
  if k = 0:
    return 0
  
  # the min is set to the max to begin with
  minBills <- k
  
  # go over every dollar value
  for i <- 1 to len(dollarValues):
    # if the current dollar value can go into k
    if k >= dollarValues[i]:
      # if the one that we check next is more than the last, we don't want to keep it
      minBills = Math.min(minBills, 1 + MinDreamDollars(k - dollarValues[i]))
  
  return minBills
  
  
MinDreamDollarsDynamicExtravaganza(k):
    MD <- array of size k
    MD[0] <- 0

    # memo time
    for i <- 1 to k:
      minBills <- k
      for j <- 1 to len(dollarValues):
        if k >= dollarValues[j]:
          # if the one that we check next is more than the last, we don't want to keep it
          minBills = Math.min(minBills, 1 + MD[i - dollarValues[i]])
        
      MD[i] = minBills
    
    return MD[k]
```

## Change Making Examples

```
ChangeCalc(k):
  if k=0
    return 0
  leastBills <- k
  for j <- each bill value <= k
    maybeLeast <- 1 + ChangeCalc(k-j)
    if maybeLeast < leastBills
      leastBills <- maybeLeast
  return leastBills
  
  
FastChangeCalc(k):
  LB[0] <- 0
  for i <- 1 through k
  leastBills <- k
    for j <- each bill value <= i
      maybeLeast <- 1 + LB[i-j]
      if maybeLeast < leastBills
        leastBills <- maybeLeast
    LB[i] <- leastBills
  return LB[k]
```

## Change Making Examples

```
bill_values = [1, 4, 7, 13, 28, 52, 91, 365]

MinDreamDollars(i):
    if i is 0:
        return 0 // no bills make 0
    minBills = i
    for each value in bill_values:
        if i >= value: // possible bill value must not be greater than desired value
            minBills = minimum of {minBills, MinDreamDollars(i-value)}
    return minBills + 1



FastMinDreamDollars(n):
    initialize MDD[0..n] to [0, 1, 2, ..., n]
    for i in 1..n:
        for each value in bill_values:
            if i >= value:
                MDD[i] = min(MDD[i],MDD[i-value]+1)
    return MDD[n]
```

## Implementation in R (just for fun)

```{r}
bills <- c(1,4,7,13,28,52,91,365)
NumBills <- function(n) {
  if(n == 0)
    return(0)
  if(n<0)
    return(Inf)
  nb <- Inf
  for(b in bills[bills <= n]) {
    tryNum <- 1 + NumBills(n-b)
    if(tryNum < nb)
      nb <- tryNum
  }
  return(nb)
}
```

Notice that you can use `Inf` for a number that is bigger than all other numbers.

## Memoized version in R


```{r}
bills <- c(1,4,7,13,28,52,91,365)
FastBills <- function(n){
  nbills <- rep(Inf, n   +1) # off by one indexing
  nbills[0  +1] <- 0
  for (i in 1:n) {
    for (b in bills[bills <= i]) {
      tryNum <- 1 + nbills[i - b   +1]
      if (tryNum < nbills[i   +1])
        nbills[i   +1] <- tryNum
    }
  }
  return(nbills[n  +1])
}
```

We can "pretend" that the `nbills` array is indexed from 1 to n by adding a "`   +1`" to every subscript (with some spaces to remind us why).

## Timing Experiments

```
> system.time(NumBills(45))
   user  system elapsed 
 23.174   0.048  23.240 
> system.time(FastBills(45))
   user  system elapsed 
  0.000   0.000   0.001 
> system.time(FastBills(455))
   user  system elapsed 
  0.001   0.000   0.001 
> FastBills(455)
[1] 5
> NumBills(455)

```

... still waiting for that last one to finish.

```
PID  USER      PR  NI    VIRT    RES    SHR S  %CPU  %MEM     TIME+ COMMAND 
2767 dhunter   20   0 1881376 217652  31936 R 100.0   1.4   3:22.98 rsession 
```

The size of the recursion tree is just silly:

```
> NumBills(4554234)
Error: C stack usage  7979108 is too close to the limit
> NumBills(4554)
Error: C stack usage  7979108 is too close to the limit
> FastBills(4554234)
[1] 12483
```

# Dynamic Programming

## Dynamic Programming Problems

- Good for problems with *recursive substructure* and *overlapping subproblems*.

## Dynamic Programming Steps

1. *Specify* the problem as a recursive function, and *solve it* recursively (with an algorithm or formula).

2. *Build the solutions* from the bottom up:
    - What are the *subproblems*?
    - What *data structure* do I store the solutions in? (Usually an array.)
    - Which subproblem does each subproblem depend on?
        - Determine an *evaluation order.*
    - Space and time?
    - Write the algorithm. Usually it will just be a couple nested loops, with the recursive calls replaced by array look-ups.

# Subset Sum

## Recall: Subset Sum

- We have a *set* $X$ of positive integers (so no duplicates).
- Given a *target* value $T$, is there some *subset* of $X$ that adds up to $T$?

## Subset Sum: Recursive Specification

- For any $x\in X$, the answer is TRUE for target $T$ if
    - The answer is TRUE for $X \setminus \{x\}$ and target $T$, or
    - The answer is TRUE for $X \setminus \{x\}$ and target $T-x$.
- In other words, the cases are:
    - *without:* There is a subset not containing $x$ that works.
    - *with:* There is a subset containing $x$ that works.
- Base Cases:
    - If $T = 0$, the answer is TRUE. ($\emptyset$ works!)
    - If $T < 0$, the answer is FALSE. (everything's positive)
    - If $X = \emptyset$, the answer if FALSE (unless $T=0$, of course.)

## Algorithm description, using indexes (revised)

```{javascript, eval=FALSE}
// X[1..n] is an array of positive integers (n is constant)

SubsetSum(i, T) :  //  Does any subset of X[i..n] sum to T?  
    if T = 0
        return TRUE
    else if T < 0 or i > n
        return FALSE
    else
        with <- SubsetSum(i + 1, T − X[i])
        wout <- SubsetSum(i + 1, T)
        return (with OR wout)

// Call this function as: SubsetSum(1, T)
```

## Subset Sum: Dynamic programming solution

- Subproblems: `SubsetSum(i,T)` depends on `SubsetSum(i+1, T')`, for $T'\leq T$.
- Data Structure: Two-dimensional *boolean* array: `SS[0..n, 0..T]`

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(252021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

## Evaluation Order?

1. Draw arrows in this diagram that the evaluation order must obey.

$$ \scriptsize
\begin{array}{ccccccccc}
       & \vdots & & \vdots && \vdots && \vdots &\\ 
\cdots & \mathtt{SS[i-1, j-1]} & \phantom{\rightarrow} & \mathtt{SS[i-1, j]} & \phantom{\rightarrow} & \mathtt{SS[i-1, j+1]} & \phantom{\rightarrow} &\mathtt{SS[i-1, j+2]} & \cdots \\[20pt]
\cdots & \mathtt{SS[i, j-1]}& & \mathtt{SS[i, j]} & \phantom{\rightarrow} & \mathtt{SS[i, j+1]} & \phantom{\rightarrow} &\mathtt{SS[i, j+2]} & \cdots\\[20pt]
\cdots & \mathtt{SS[i+1, j-1]} & & \mathtt{SS[i+1, j]} & & \mathtt{SS[i+1, j+1]} & &\mathtt{SS[i+1, j+2]} & \cdots \\[20pt]
\cdots & \mathtt{SS[i+2, j-1]}  & & \mathtt{SS[i+2, j]} & & \mathtt{SS[i+2, j+1]} & & \mathtt{SS[i+2, j+2]} & \cdots \\
       & \vdots & & \vdots && \vdots && \vdots &\\ 
\end{array}
$$



## Describe the algorithm

3. Propose an evaluation order, and write some for-loops. 

4. Space and time?

5. If time permits, complete the memoized algorithm.

#  Subset Sum postmortem

## Subset Sum: Memoized version (spoiler)

```{javascript, eval=FALSE}
// X[1..n] is an array of positive integers (n is constant)
// SS[1..(n+1), 0..T] is a boolean array

FastSubsetSum(T):
    Initialize SS[*, 0] to TRUE
    Initialize SS[n + 1, *] to FALSE for * > 0
    Initialize SS[* , *] to FALSE for * < 0    // hmmm...
    for i <- n downto 1
        for t <- 1 to T
            with <- SS[i + 1, t - X[i]]
            wout <- SS[i + 1, t]
            SS[i, t] <- (with OR wout)
    return S[1, T]
```


## Spiffier Version (avoid negative indexes)

```{javascript, eval=FALSE}
// X[1..n] is an array of positive integers (n is constant)

FastSubsetSum(T):
    Initialize SS[*, 0] to TRUE
    Initialize SS[n + 1, *] to FALSE for * > 0
    for i <- n downto 1
        for t <- 1 to X[i] − 1
            SS[i, t] <- SS[i + 1, t]   // avoid t - X[i] < 0 case
        for t <- X [i] to T
            SS[i, t] <- SS[i + 1, t] OR SS[i + 1, t − X[i]]
    return S[1, T]
```

- Space and Time are both $O(nT)$.


## Not so fast!

- $O(nT)$ seems like an improvement over $O(2^n)$.
    - It is better for small $T$.
- However, $O(nT)$ is bad when $T$ is large.
    - If $m$ is the number of *bits* in $T$, then $T \in O(2^m)$, so the space and time become $O(n\cdot 2^m)$.
    - So in terms of the *size* of the input, this algorithm is still exponential.
    
Fact: The SubsetSum problem is NP-complete.
    