---
title: "Test #1 Review"
date: "February 8, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## Writing Recursive Functions

> Your **only** task is to simplify the original problem, or to solve it directly when simplification is either unnecessary or impossible; the *Recursion Fairy* will solve all the simpler subproblems for you, using Methods That Are None Of Your Business...

The Recursion Fairy is the inductive hypothesis.

## Candy Swap Saga

- If you swap your candy for another candy of the same type, you earn
one point.
- If you swap your candy for a candy of a different type, you lose one point.
- If you visit an animal and decide not to swap candy, your score does not
change.

Would the greedy algorithm work?

## Recursive Specification in English

Find the highest possible score by comparing:

- The highest score for the rest of the animals after choosing to swap,  versus 
- The highest score for the rest of the animals after choosing not to swap.

To make the decision for `animal[i]`, let the Recursion Fairy give you the maximum score for the rest of the animals. (No for-loop needed.)

## Recursive Description

```{javascript, eval=FALSE}
CandySwap(i,h):
  if h=C[i]:
    point <- 1
    if i=n:
      return 1
  else:
    point <- -1
    if i=n: 
      return 0
  else:
    swap <- point + CandySwap(i+1, C[i])
    skip <- CandySwap(i+1, h)
    return max(swap, skip)
```

## Dependencies, data structure, evaluation order

- At any point in the recursive problem, the subproblems are CandySwap(i+1, held_candy) and CandySwap(i+1, animals[i]). 

- We can memoize past results in a 2-dimensional array CS based on the two input values. The first value we can fill in is CS[n+1][1..3] = 0, and from there we can go in descending order with n, n-1, n-2, et cetera.

## Dynamic Programming Solution

```{javascript, eval=FALSE}
CandySwapMemo(h):
  initialize CS[n, *] to 0 for 0 < * < 4
  CS[n,C[n]] <- 1   
  for i <- (n-1) down to 1:   
    for j <- 1 to 3:   
      if C[i]=j:   
        point <- 1   
      else: 
        point <- -1   
      swap <- point + CS[i+1, C[i]]   
      skip <- CS[i+1,j]
      CS[i,j] <- max(swap, skip)
  return CC[1, h]
```

Time: $O(n)$, Space: $O(n)$

## Can improve space

Note: you can make it constant space, since the solution only depends on the *previous* subproblem

```{javascript, eval=FALSE}
C[1..n]       // the types of candy held by each animal
PrevMSC[1..3] // memoization structure for max scores of candy, depending on current candy held
tmp[1..3]     // a temporary array for storing max scores of candy
FastMaxScoreCandy(CurrentCandy):
    PrevMSC[C[n]] <- 1                    // we can get a point in the end if we swap
    PrevMSC[all candies except C[n]] <- 0 // or maybe we stick with what we have
    from animal <- n-1 down to 1:
        for candy <- 1..3:
            if candy is C[animal]:
                DoSwapScore <- 1 + PrevMSC[C[animal]]
            otherwise
                DoSwapScore <- -1 + PrevMSC[C[animal]]
            DontSwapScore <- PrevMSC[candy]
            tmp[candy] <- max(DoSwapScore, DontSwapScore)
        copy tmp -> PrevMSC
    return PrevMSC[CurrentCandy]
```

## Candy Swap Saga: Recursive Solution in R

```{r}
set.seed(4323)
candyTypes <- c("peanut", "heath", "truffle")
n <- 20
animalCandy <- sample(candyTypes, n, replace=TRUE)

maxScore <- function(i, myCandy) { # max score from visiting animals i..n holding myCandy
  if(i > n)
    return(0)
  if(myCandy == animalCandy[i])
    swap <- 1 + maxScore(i+1, myCandy)
  else
    swap <- -1 + maxScore(i+1, animalCandy[i])
  same <- maxScore(i+1, myCandy)
  return(max(swap, same))
}
```

## Candy Swap Saga: Iterative Solution in R

```{r}
FastMaxScore <- function() {
  S <- matrix(rep(0, 3*(n+1)), nrow = n+1, ncol = 3)
  for(i in n:1) {
    currentAnimalCandyNum <- which(candyTypes == animalCandy[i])
    # holding peanut: Score goes in S[i, 1]
    # holding heath: Score goes in S[i, 2]
    # holding truffle: Score goes in S[i, 3]
    for(myC in 1:3) {
      if(candyTypes[myC] == animalCandy[i])
        swap <- 1 + S[i+1,myC]
      else
        swap <- -1 + S[i+1,currentAnimalCandyNum]
      same <- S[i+1, myC]
      S[i, myC] <- max(swap, same)
    }
  }
  return(S[1,1])
}
```

## Tests

```{r}
n <- 4
animalCandy <- c("heath", "heath", "heath", "heath")
maxScore(1, "peanut")
FastMaxScore()

set.seed(4323)
candyTypes <- c("peanut", "heath", "truffle")
n <- 20
animalCandy <- sample(candyTypes, n, replace=TRUE)
FastMaxScore()
```

## Timing Experiments

```{r}
system.time(maxScore(1, "peanut"))
system.time(FastMaxScore())
```

# Test #1 Contents

## Study Guide on Canvas

- Questions on the [Study Guide](https://westmont.instructure.com/courses/5960/assignments/55827)?
- See also the [slide summary](../slide01-11notes.html).

# Practice Exercises

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(282021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

