---
title: "Greedy Algorithms: Scheduling"
date: "February 12, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Greedy Algorithms

## The Greedy Choice

- Problem: Find the optimal (e.g., cheapest/best) long-term sequence of choices.
- A *greedy algorithm* makes the cheapest/best short-term choice at each stage of the process.

We saw some examples where the greedy choice *doesn't* work:

- Change-making in Nadirian currency.
- Candy Swap Saga

## Greedy vs. Dynamic Programming

- Both solve *optimization* problems.
- Problems have *recursive substructure.*
   - Make a choice, then optimize the remainder.
- There is one **key difference:**
    - In the dynamic programming solutions, we had to try all the different choices at each stage.
        - Churro cutting: Try all the different cuts, then optimize the rest.
        - Candy swap: Try swapping and not swapping, then optimize the rest.
    - A greedy algorithm just makes the optimal *short-term* choice.
        - Cut off the most expensive piece.
        - Never swap for a different candy.
   
## Isn't this easy?

- Writing greedy algorithms is easy. (And they're usually fast.)
- The hard part is **proving that they actually give you an optimal solution.**

# Scheduling

## Maximize the number of events

- You have a list of potential events to schedule (e.g., Fringe Festival skits.)
   - Each event has a starting time $S[i]$ and a finish time $F[i]$.
   - You only have one stage. 

- The organizer of the festival only cares about *maximizing the number* of events.
- Which events should you choose?

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(2122021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

## How many can you schedule?

```{r, echo=FALSE}
S <- c(5,1,6,1,4,9 ,2,4)
F <- c(7,5,9,6,7,10,4,6)
rbind(S,F)
```

Find the maximum number of events for this input. (Draw a picture?)

# Greedy Scheduling Algorithm

## First: Recursive Structure?

```{javascript, eval=FALSE}
// There are n events, numbered 1..n
// S[1..n] is a list of starting times
// F[1..n] is a corresponding list of finishing times

MaxNumEvents(X)  // the maximum number of events in X ⊆ {1..n} that can be scheduled 
    if X = ∅ then
        return 0
    Choose any element k in X
    B <- the set of events in X ending BEFORE S[k]
    A <- the set of events in X starting AFTER F[k]
    return the maximum of:
        MaxNumEvents(X \ {k}) // do not use event number k
        MaxNumEvents(A) + 1 + MaxNumEvents(B)  // use event k
```

It works, by induction.

## Greedy Scheduler

```{javascript, eval=FALSE}
// There are n events, numbered 1..n
// S[1..n] is a list of starting times
// F[1..n] is a corresponding list of finishing times

MaxNumEvents(X)  // the maximum number of events in X ⊆ {1..n} that can be scheduled 
    if X = ∅ then
        return 0
    Let k in X be the event that ends first // GREEDY CHOICE
    A <- the set of events in X starting AFTER F[k]
    return MaxNumEvents(A) + 1
```

- Summary: Choose the event $k$ that *ends first*, discard events that conflict with $k$, and recurse.
- How do we know that this works?

# Greedy Proof of Correctness

## Greedy choice property: Idea of proof

To prove that a greedy algorithm works, you need to show that it has the *greedy choice property*.

- Suppose you have a non-greedy (optimal) solution.
- Show that the greedy solution is just as good.
    - Find the first place where they differ, and show that you could have used the greedy choice without making the solution less optimal.
    
## Group Exercise: Find another solution

Suppose the sequence of events continues, but we don't have all the values:

```
S <- [5, 1, 6, 1, 4, 9, 2, 4  ... ]
F <- [7, 5, 9, 6, 7,10, 4, 6, ... ]
```

- Your greedy solution contains 17 events, starting with: 7, 8, 3, 6, ...

- Buford has found another 17-event solution: 7, 8, 13, 11, ...

> Claim: If you replace the 13 in Buford's solution with a 3, you get yet another 17-event solution.

*Write a couple sentences* to justify this claim.

## Greedy Scheduler Correctness

```{javascript, eval=FALSE}
// There are n events, numbered 1..n
// S[1..n] is a list of starting times
// F[1..n] is a corresponding list of finishing times

MaxNumEvents(X)  // the maximum number of events in X ⊆ {1..n} that can be scheduled 
    if X = ∅ then
        return 0
    Let k in X be the event that ends first // GREEDY CHOICE
    A <- the set of events in X starting AFTER F[k]
    return MaxNumEvents(A) + 1
```

- Suppose that `MaxNumEvents({1..n})` returns $m$.
- Let $g_1, g_2, \ldots, g_m$ be the events chosen by this greedy algorithm.
- Suppose that $g_1, g_2, \ldots, g_{j-1}, \mathbf{c_j}, c_{j+1}, \ldots, c_m$ is another sequence of compatible events.
    - Same length $m$, but different choice for event number $j$.
    
Now apply the same reasoning as in the Buford example.

## Proof of correctness

**Proof.**  Suppose that `MaxNumEvents({1..n})` returns $m$. Let $g_1, g_2, \ldots, g_m$ be the events chosen by this greedy algorithm. Suppose that 
$$g_1, g_2, \ldots, g_{j-1}, \mathbf{c_j}, c_{j+1}, \ldots, c_{m'}$$ 
is another sequence of compatible events, with $m' \geq m$.

Since event $g_j$ is part of the first solution, it must start after event $g_{j-1}$. Since event $g_j$ was chosen by the greedy algorithm, *it must end before* all the events in $X \setminus \{g_1, g_2, \ldots, g_{j-1}\}$. In particular, it must end before event $c_{j+1}$. Therefore, we can replace $c_j$ with $g_j$ and obtain another solution:
$$g_1, g_2, \ldots, g_{j-1}, \mathbf{g_j}, c_{j+1}, \ldots, c_{m'}$$ 
Continuing in this way, all of the $c_i$'s can be replaced with $g_i$'s. Therefore $m = m'$, and the greedy solution is optimal.

# Refining the algorithm

## Greedy Scheduler Time

```{javascript, eval=FALSE}
// There are n events, numbered 1..n
// S[1..n] is a list of starting times
// F[1..n] is a corresponding list of finishing times

MaxNumEvents(X)  // the maximum number of events in X ⊆ {1..n} that can be scheduled 
    if X = ∅ then
        return 0
    Let k in X be the event that ends first // GREEDY CHOICE
    A <- the set of events in X starting AFTER F[k]
    return MaxNumEvents(A) + 1
```

- Non-recursive work: $O(n)$
    - Making the greedy choice requires scanning $X$, which is $O(n)$.
    - Forming $A$ is also $O(n)$
- The function uses *tail recursion*, so it could be written as a loop.
    - The number of remaining events decreases by at least 1 each time, so this "loop" runs $O(n)$ times. **Total time:** $O(n^2)$

## Improve using an efficient sort

```{javascript, eval=FALSE}
MaxNumEvents(S[1..n], F[1 .. n]):
    sort F and permute S to match  // O(n log(n))
    count <- 1
    X[count] <- 1   // Because now event 1 has first finish time. 
    for i <- 2 to n  // Scan events in order of finish time. 
        if S[i] > F [X[count]]  // Is event i compatible?
            count <- count + 1  //    If so, it is the
            X[count] <- i       //    greedy choice.
    return count
```

Time: 

- $O(n \log n)$ to sort (e.g., use MergeSort)
- After that finishes, there's just a linear time for-loop: $O(n)$.
- **Total time:** $O(n \log n)$ 

