---
title: "Binary Character Codes"
date: "February 17, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## Interval Stabbing

## Easiest Greedy Choice

Choose the interval with the *smallest right endpoint*, stab at this right endpoint, eliminate the intervals that have been stabbed, and recurse.

(picture)

## Exchange Argument

- Let $g_1, g_2,\ldots,g_m$ be the stabs made by the greedy algorithm (in order from left to right), and suppose $g_1,g_2,\ldots, g_{j-1}, c_j, c_{j+1},\ldots, c_{m'}$ is another set of stabs (in order) that stab all the intervals, with $m' \leq m$.

- $c_j$ must stab the interval $I$ with the leftmost right endpoint, after eliminating those stabbed by $g_1,g_2,\ldots, g_{j-1}$.

- Moving $c_j$ to $g_j$ (the right endpoint of $I$) stabs all of the same intervals, because no other intervals end before interval $I$ does.

- In this way, all of the $c_i$'s can be replaced with $g_i$'s, so $m = m'$.

## Recursive Implementation

```
Algorithm(L[1..n], R[1..n])
if n = 0 //empty list
  return 0
else
  min = infinity
  for i <- 1 to n
    if R[i] < min
      min = R[i] //finding stabpoint
  newL = L[1..n]
  newR = R[1..n]
  for j <- 1 to n
    if min >= L[j] && min <= R[j] //finding if stabpoint overlaps with each brick
      newL[] <- newL \ L[j] //takes away that brick from array
      newR[] <- newR \ R[j] 
  return 1 + Algorithm(newL[], newR[])
```

Time: $O(n^2)$ (linear work inside tail recursion)

## Iterative Implementation

```
Stabby(R[1..n], L[1..n]):
  counter <- 0 
  MergeSort R and permute L to match 
  i <- 1
  while i <= n 
    stab <- R[i]
    counter++
    while L[i] <= stab  \\ then it got stabbed
      i++  \\ keep checking and skipping what gets stabbed
  return counter
```

Time: $O(n \log n)$, because the loop runs in $O(n)$ time, incrementing $i$ until it exceeds $n$.

**Note:** The second while loop won't remove intervals that have later end times than some unstabbed interval. But these will get removed eventually; at the very latest, at the last stab. Such an interval would never have the leftmost right endpoint, and so would never determine the location of a stab.  So it won't affect the greedy choices, or the final value of `counter`.

# Encoding Characters as Binary Strings

## Codes

- A **binary code** is a list of *code words*, which are binary strings.
- Given an *alphabet* of characters, a **binary character code** is a binary code where each character corresponds to a code word.
    - e.g., ASCII, UTF-8

Example:

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
code word | 000 | 001 | 010 | 011 | 100 | 101 |
```

- Given a character string, can *encode* it into binary.
    - $\mathtt{YO} \mapsto \mathtt{101011}$
- Given a binary string, can *decode* it to a character string.
    - $\mathtt{011100010} \mapsto \mathtt{OUI}$

## Compression

- Goal: Minimize number of bits needed to encode.
- Idea: Use code words of different lengths.
    - Problem: How to parse a binary string so you know where the code words start?
    
Example: 

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
code word |  00 |  01 |  10 |  11 | 100 | 101 |
```

1. Find two different decodings of `10110010`.

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(2172021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

# Prefix Codes

## Unambiguous encoding

- In a **prefix code**, no code word can start with another code word.
    - i.e., no code word can be a *prefix* for some other code word.

Example:

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
code word |  10 | 111 | 001 | 000 | 110 |  01 |
```

Now there's only one way to decode anything:
$$
\mathtt{101100110} \mapsto \mathtt{AUYA}
$$

## Prefix Codes and Trees

- A binary tree whose *leaves are the characters* determines a prefix code.
- Left edges correspond to 0s. Right edges correspond to 1s.

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
code word |  10 | 111 | 001 | 000 | 110 |  01 |


                             #
                        0/       \1
                        #         #
                     0/   \1   0/   \1
                     #     Y   A     #
                   0/ \1           0/ \1
                   O   I           U   E 
```

# Optimal Prefix Codes

## Optimize the number of bits in a message

Suppose we want to encode the following message:

```{r, echo=FALSE}
mess <- c(rep("E", 80), rep("A", 50), rep("O", 30), rep("I", 30), rep("U", 20), rep("Y", 20))
cat(paste(sample(mess, length(mess)), collapse=""))
```

This message contains 80 Es, 50As, 30 Os, 30Is, 20Us, and 20Ys.

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
frequency |  50 |  80 |  30 |  30 |  20 |  20 |
```

We want to construct a prefix code that uses the fewest number of bits for the whole message.

## Can you do better?

The message contains 80 Es, 50As, 30 Os, 30Is, 20Us, and 20Ys.

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
frequency |  50 |  80 |  30 |  30 |  20 |  20 |
```

Encode it using the following prefix code:

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
code word |  10 | 111 | 001 | 000 | 110 |  01 |
```

This requires `50*2 + 80*3 + 30*3 + 30*3 + 20*3 + 20*2` = `r 50*2 + 80*3 + 30*3 + 30*3 + 20*3 + 20*2` bits.

2. Why?
3. Propose a prefix code that is able to encode the message using fewer bits. How many bits can you save?

# Huffman Encoding

## Huffman's Greedy Algorithm

**Given:** A list of characters `C[1..n]` and their frequencies `F[1..n]` in the message we want to encode.

**Goal:** Construct a binary tree for a prefix code that uses the fewest number of bits to encode the message.

**Algorithm:** Start with a "forest" of leaves: every character is a leaf.

- Find the two trees in this forest with smallest total frequency.
- Join these two trees under a common root.
- Recurse, until the forest has just one tree.

## Huffman's Algorithm: Example

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
frequency |  50 |  80 |  30 |  30 |  20 |  20 |
```

Start with a "forest" of leaves: every character is a leaf:

```
A:50    E:80    I:30    O:30    U:20    Y:20
```

Find the two trees in this forest with smallest total frequency. Join these two trees under a common root:

```
A:50    E:80    I:30    O:30      #:40
                                 /   \
                                U     Y
```

Recurse:

```
A:50    E:80         #:60          #:40
                   /   \         /   \
                  I     O       U     Y
```

## Huffman's Algorithm: Example, continued

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
frequency |  50 |  80 |  30 |  30 |  20 |  20 |
```

```
A:50    E:80         #:60          #:40
                   /   \         /   \
                  I     O       U     Y
```

Recurse:

```
 E:80                #:60          #:90
                   /   \         /   \
                  I     O       A     #
                                    /   \
                                   U     Y
```

Recurse:

```
                     #:140         #:90
                   /   \         /   \
                  E     #       A     #
                       / \          /   \
                      I   O        U     Y
```

## Huffman's Algorithm: Optimal Code?

```
                     #:140         #:90
                   /   \         /   \
                  E     #       A     #
                       / \          /   \
                      I   O        U     Y
```

One final recursion:

```
                            #
                       /         \
                     #             #
                   /   \         /   \
                  E     #       A     #
                       / \          /   \
                      I   O        U     Y
```

Optimal code?

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
code word |  10 |  00 | 010 | 011 | 110 | 111 |
```

## Sizes of code words?

4. Apply Huffman's algorithm to produce a prefix code, given the following frequencies. Does this algorithm minimize the length of the longest code word?

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
frequency |  50 | 100 |  30 |  30 |  20 |  20 |
```

5. Find an assignment of frequencies to characters in order to produce the widest possible assortment of code word lengths using Huffman's algorithm. Is it possible to produce a code where all the code words have different lengths?

```
character |  A  |  E  |  I  |  O  |  U  |  Y  |
frequency |     |     |     |     |     |     |
```




