---
title: "Huffman Codes"
date: "February 19, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## Prefix Codes

```
| letter   |   b   |   l   |   a   |   c   |   k   |   h    |   w   |   s   |
| codeword |   01  |   11  |  0010 |  0001 |  0011 |  0000  |  010  |  011  |
```

Since 01 is a prefix for more than one letter (b,s,w) this is not a prefix code.


```
                             #
                        0/       \1
                        #         e
                     0/   \1        
                     a     #        
                         0/ \1     
                         #   g
                       0/ \1
                       m   n      
```

- 10110110100000101 ---> `eggman`
- This is not an optimal code because the letter `g` was used the most but it did not use the least amount of code to be encoded to: `g` and `e` switched in the binary tree would have used fewer bits.

## Huffman's algorithm

Notice that the process is not uniquely determined, because sometimes there are multiple options with the same frequencies.

```
a:100       b:20        c:30        d:20        e:150       f:10        g:20        h:40     i:110

                                                
a:100       b:20        c:30        e:150       #:30        g:20        h:40        i:110
                                                /   \
                                               d     f        

a:100       c:30        e:150       #:30        #:40        h:40        i:110
                                    /   \      /   \
                                   d     f    b     g

a:100       e:150       #:60        #:40        h:40        i:110
                        /   \       /  \
                       c     #     b    g
                            / \
                           d   f     

a:100       e:150       #:60        #:80        i:110
                        /   \       /   \
                       c     #     h     #
                            / \         / \
                           d   f       b   g
                           
a:100       e:150               #:140           i:110
                                /   \
                           #           #
                         /   \       /   \
                       c     #     h     #
                            / \         / \
                           d   f       b   g

e:150       #:210               #:140
            /   \               /   \
           a     i         #           #
                         /   \       /   \
                       c     #     h     #
                            / \         / \
                           d   f       b   g   

#:210         #:290
 /  \       /      \
a    i     e        #
               /      \
             #         # 
           /   \     /  \
          c     #   h    #  
              /   \     /  \
             d     f   b    g
             
      #:500
    /       \
   #            #
 /  \       /      \
a    i     e        #
               /      \
             #         # 
           /   \     /  \
          c     #   h    #  
              /   \     /  \
             d     f   b    g
```

## Internal nodes

You can think of internal nodes as being "combo" characters.

```
             aifbcdghe
           /0         1\
         ai             fbcdghe
       /0  1\          /0      1\
      a      i    fbcdgh         e
                /0      1\
             fbc          dgh
            /0  1\       /0 1\
          fb      c    dg     h
        /0  1\       /0  1\
       f      b     d      g
```

## Priority Queues

- **Max Heap:** For every node, the value of its parent must be greater than its own value. To fix this tree, we must swap 16 and 17.

```
              24
            /    \
           21    17
          / \    / 
         9  18  16
```

- `ExtractMax`, `ExtractMin`, `Insert`, are all $O(\log n)$.
    - `Peek` is $O(1)$, because you can just read the root value and not change anything.

## {data-background="https://media.giphy.com/media/T3fwN6Pbm3ZPa/giphy.gif" data-background-size="contain"}

## Building a Priority Queue

- Adding elements to the heap takes $O(\log n)$ for each element, but we are doing this for $n$ elements, so we multiply the times together to get $O(n\log n)$.
    - This is an upper bound, but it is *not tight*.
- There are at most $\left\lceil n/2^{h+1} \right\rceil$ nodes of height $h$.

$$
\sum_{h=0}^{\lfloor \log_2 n \rfloor} \left\lceil \frac{n}{2^{h+1}}\right\rceil O(h) = O\left( n \sum_{h=0}^{\lfloor \log_2 n \rfloor}\frac{h}{2^{h+1}} \right)= O\left( n \sum_{h=0}^{\infty}\frac{h}{2^{h+1}} \right) = O(n)
$$

- So it turns out that $O(n)$ is a **tight** upper bound on the time needed to build a priority queue.



# Huffman Algorithm Correctness

## Part 1: Exchange Argument

Given a list of characters `C[1..n]` and their frequencies `F[1..n]` in the message we want to encode.

- Let $T_G$ be a tree produced by Huffman's greedy algorithm.
- Suppose $T$ is some other tree giving an optimal prefix code.
    - Let $x$ and $y$ be the characters with smallest frequency.
    - Let $a$ and $b$ be the leaves of the lowest subtree of of $T$.
        - If $a$ and $b$ are $x$ and $y$, we're done.
        - Otherwise, swap $a$ with $x$ and $b$ with $y$, and you get a code that is at least as efficient.
        
**Upshot:** The first greedy choice is optimal.

## Part 2: Structural Inductive Argument (sketch)

- Let $C'$ be the character set $C$ with the least frequent characters $x$ and $y$ replaced with the character $x\!\!\!y$. 
- Suppose (inductive hypothesis) $T'$ is an optimal tree for $C'$ that was built by Huffman's algorithm.
    - $x\!\!\!y$ is a leaf of $T'$.
    - Replace this leaf with: $\begin{array}{ccc} & /\backslash & \ \\  & x \;\;\; y&  \end{array}$
    - cost of $T$ = cost of $T'$ + frequencies of $x$ and $y$. (details in book)
    - Since $T'$ had minimal cost, so must $T$.

## Full Binary Trees

- Usual recursive definition: 
    - A *full binary tree* is made from two full binary trees joined under new root.
- Alternative recursive definition: 
    - A *full binary tree* is made by replacing a leaf of a full binary tree with a two-leaf subtree.
    
Either way, a full binary tree with $n$ leaves has $2n-1$ nodes. (picture)

# Huffman Algorithm Implementation

```{javascript, eval=FALSE}
// indices 1 .. n are the leaves (corresponding to C[1..n])
// index 2n-1 is the root 
// P[i] = parent of node i
// L[i] = left child of node i, R[i] = right child
BuildHuffman(F[1..n]):
    for i <- 1 to n
        L[i] <- 0; R[i] <- 0
        Insert(i, F[i]) // put indices in priority queue, frequency = priority
    for i <- n + 1 to 2n − 1  // ??typo in book?? (Book starts this loop at n, not n+1.)
        x <- ExtractMin()     // extract smallest trees in forest
        y <- ExtractMin()     //   from the priority queue
        F[i] <- F[x] + F[y]   // new internal node
        Insert(i, F[i])       // put new node into priority queue
        L[i] <- x; P[x] <- i  // update children
        R[i] <- y; P[y] <- i  //    and parents
    P[2n − 1] <- 0            // root has no parent
```

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(2192021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

## Explore Huffman's Algorithm

1. Find the running time of this implementation of Huffman's Algorithm.

2. Construct the arrays F, P, L and R for the following table of frequencies. In other words, `F[1..6] = [50, 80, 30, 30, 20, 20]`.

```
            character |  A  |  E  |  I  |  O  |  U  |  Y  |
            frequency |  50 |  80 |  30 |  30 |  20 |  20 |
```
    
3. Am I right that there's a typo in the book in the limits on the second for-loop??


