---
title: "Trees"
date: "March 12, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Trees

## Applications of Trees

- Binary search trees
- Recursion trees
- DFS and BFS trees
- Binary heaps
- Huffman Codes
- etc.

Most of these trees had a preferred *root*, and a preferred *direction* (away from the root).

## Tree: General Definition

> A **tree** is a connected, acyclic graph.

By default, a generic tree is *undirected*. To understand this definition, we need to be precise about the meaning of *acyclic*.

- A **walk** in an undirected graph $G$ is a sequence of vertices, where each
adjacent pair of vertices are adjacent in $G$.
- A walk is **closed** if it starts and ends at the same vertex. 
- A walk is called a **path** if it visits each vertex at most once.
- A **cycle** is a closed walk that enters and leaves each vertex at most once.
- An undirected graph is **acyclic** if no subgraph is a cycle.

## Unravel the definitions

1. Draw an example of a walk that is not a path.

2. Draw an example of a closed walk that is not a cycle.

3. What is a closed path? Can you draw an example?

4. Draw a tree with 6 vertices. How many edges does it have?

5. How many edges does a tree with $n$ vertices have? Can you prove your answer?

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(3122021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

# Paths in trees

## Unique Paths

**Lemma.** Suppose $T$ is a tree, and $u$ and $v$ are two vertices in $T$. Then there is a unique path from $u$ to $v$.

- There exists a path from $u$ to $v$ because $T$ is connected.
- Suppose to the contrary that there are two different paths from $u$ to $v$.
    - Path 1: $a_0 = u, a_1, a_2, \ldots, a_k = v$.
    - Path 2: $b_0 = u, b_1, b_2, \ldots, b_j = v$.
        - Let $a_s = b_s$ be the last vertex at the start of these paths that are the same.
        - Let $a_{k-e} = b_{j-e}$ be the first vertex at the end of these paths that are the same.
    - Then $a_s, \ldots, a_{k-e}, b_{j-e-1}, \ldots, b_s$ is a cycle, contradicting that $T$ is a tree.

## Equivalent Condition

The converse of the Lemma is also true: If $G$ is a graph with the property that every pair of vertices are joined by a unique path, then $G$ is a tree.

- Connected: $\checkmark$
- Acyclic: A cycle would contradict the unique path condition.

# Weighted Graphs

## Weighted Edges

A graph is called *weighted* if there is an assignment of real numbers (weights) to the edges.  For example,

- Cost function in a network.
- Distance function in a spatial arrangement.

## Minimum Spanning Tree of a Weighted Graph

Given a weighted graph $G$, a **minimum spanning tree** $T$ is a subgraph of $G$ such that:

- $T$ contains all the vertices of $G$, and
- The sum of weights of the edges of $T$ is as small as possible.

6.  Why do the above two conditions guarantee that $T$ is a tree?

7. Draw an example of a weighted graph with two different minimum spanning trees (if possible).

8. Draw an example of a weighted graph with a unique smallest edge $e$, such that the minimum spanning tree does not contain $e$ (if possible).

# Uniqueness

## Minimum spanning is unique (almost)

**Lemma.** If all edge weights in a connected graph $G$ are distinct, then $G$ has a unique minimum spanning tree.

- Suppose $G$ has two minimum spanning trees $T$ and $T'$.
- Let $e$ be the smallest edge in $(T\setminus T')\cup(T'\setminus T)$.
    - WLOG, $e\in T\setminus T'$.
- Adding this edge to $T'$ creates a cycle.
    - One of the edges in this cycle is not in $T$.
    - Exchange this edge for $e$, and you reduce the weight of $T'$, a contradiction.
