---
title: "Borůvka's MST Algorithm"
date: "March 24, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Assignment Comments, Sample Solutions

## MST with largest edge

1. Give an example of a weighted undirected graph that is not a tree, but whose minimum spanning tree contains the largest edge in the graph.

```
         1.1
       0--------0
   6.7/ \        \2.4
     /   \5.2     0
    0     \       /2.3
           0-----0
             3.3
```


## Largest cycle edge excluded

2. Let $G$ be a weighted, undirected graph. Prove that for any cycle in $G$, the minimum spanning tree *does not include* the largest edge in the cycle.

(Common wrong answer on Jamboard.)

## Proof considering components (sketch)

- Suppose there exists a minimum spanning tree $T$ that includes a maximum edge $e_\text{max} = uv$ of  some cycle. Remove $e_\text{max}$ from $T$. 

- The tree is no longer a minimum spanning tree but instead two components, one containing $u$ and the other containing $v$. But since there was a cycle in $G$ containing $e_{max}$, there is a path in $G$ from $u$ to $v$. Therefore, some edge $e$ on this path must connect the two components.

- Exchange $e$ and $e_\text{max}$, contradicting minimality.

## Disprove with a counterexample

3. Prove or disprove: The minimum spanning tree of a weighted, undirected graph $G$ includes the smallest edge in *every* cycle in $G$.

```
     (10)     (3)      (11)
e --------- a --- b ----------- d
 \           \   /             /
  \_      (2) \ / (1)        _/
    \          c            /
     \_                   _/
       \                 /
   (16) \_             _/ (15)
          \           /
           \_       _/
             \     /
              \_ _/
                f
```

# Safe and Useless

## Minimum Spanning Tree of a Weighted Graph

- Given a weighted graph $G$, a **minimum spanning tree** $T$ is a subgraph of $G$ such that:
    - $T$ contains all the vertices of $G$, and
    - The sum of weights of the edges of $T$ is as small as possible.
- An *intermediate spanning forest* is a subgraph of the MST.

For any intermediate spanning forest $F$ of a graph $G$,

- An edge of $G$ is **useless** if it is not an edge of $F$, but both its endpoints are in the same component of $F$.
- An edge of $G$ is **safe** if, in some component of $F$, it is the minimum-weight edge with exactly one endpoint in the component.

## Prim's Theorem

**Theorem.** Let $F$ be an intermediate spanning forest of a weighted graph $G$, and suppose $e$ is a safe edge. Then the minimum spanning tree contains $e$.

>- Let $C$ be the component of $F$ that $e$ is "safe" for, so $e$ has an endpoint in $C$ and an endpoint in $G \setminus C$, and $e$ is the smallest such edge.
>- Suppose to the contrary that the minimum spanning tree $T$ does not contain $e$. The endpoints $u$ and $v$ of $e$ must be connected by a path in $T$, and since exactly one of the endpoints of $e$ is in $C$, one of the edges $e'$ of this path must have an endpoint in $C$ and an endpoint not in $C$. 
>- Removing $e'$ from $T$ creates a spanning forest with two components, one containing $u$ and another containing $v$. Adding $e$ to this forest connects these components, and so creates a tree. But this new tree has smaller total weight than $T$, because $e$ is smaller than $e'$. That's a contradiction.

## Greedy MST Meta-Algorithm

Let $G$ be a weighted graph with vertex set $V$ and edge set $E$.

```{javascript, eval=FALSE}
MetaMST(V,E)
    F = (V, ∅)
    while F is not a tree
        add a safe edge to F   // NOTE: This choice needs to be specified!!
    return F
```

(See Jamboard from last time.)

# Borůvka's Algorithm for the minimum spanning tree

## General idea

Use `MetaMST`, but add *all* the safe edges inside the `while`-loop.

```{javascript, eval=FALSE}
MetaMST(V,E)
    F = (V, ∅)
    while F is not a tree
        add a safe edge to F   // NOTE: This choice needs to be specified!!
    return F
```

In order to analyze the time, we need to specify how the safe edges are located.

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(3242021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

## Label the vertices with their component number

```{javascript, eval=FALSE}
CountAndLabel(G):                         LabelOneComponent(s,count):
    count <- 0                                Enqueue(s)
    for all vertices v                        while the queue is not empty
        unmark v                                  v <- Dequeue
    for all vertices v                            if v is unmarked
        if v is unmarked                              mark v               // Line #1
            count <- count + 1                        v.comp <- count
            LabelOneComponent(v, count)               for each edge vw 
    return count                                          Enqueue(w)       // Line #2
```

Suppose $G$ has $V$ vertices and $E$ edges.

1. What is another name for the `LabelOneComponent` algorithm?

2. How many times does `CountAndLabel` execute Line #1?

3. How many times does `CountAndLabel` execute Line #2?

4. What is the running time of `CountAndLabel`?

## Borůvka's Algorithm

```{javascript, eval=FALSE}
Borůvka(V, E):                              AddAllSafeEdges(E, F, count):
    F = (V, ∅)                                  for i <- 1 to count
    count <- CountAndLabel(F)                       safe[i] <- Null
    while count > 1                             for each edge uv ∈ E
        AddAllSafeEdges(E, F, count)                if comp(u) != comp(v)
        count <- CountAndLabel(F)                       if safe[comp(u)] = Null or w(uv) < w(safe[comp(u)])
    return F                                                safe[comp(u)] <- uv
                                                        if safe[comp(v)] = Null or w(uv) < w(safe[comp(v)])
                                                            safe[comp(v)] <- uv
                                                for i <- 1 to count
                                                    add safe[i] to F   
```

## Borůvka's Algorithm

```{javascript, eval=FALSE}
Borůvka(V, E):                              AddAllSafeEdges(E, F, count):
    F = (V, ∅)                                  for i <- 1 to count
    count <- CountAndLabel(F)                       safe[i] <- Null
    while count > 1                             for each edge uv ∈ E
        AddAllSafeEdges(E, F, count)                if comp(u) != comp(v)
        count <- CountAndLabel(F)                       if safe[comp(u)] = Null or w(uv) < w(safe[comp(u)])
    return F                                                safe[comp(u)] <- uv
                                                        if safe[comp(v)] = Null or w(uv) < w(safe[comp(v)])
                                                            safe[comp(v)] <- uv
                                                for i <- 1 to count
                                                    add safe[i] to F   
```

5. What is the running time of the *first* `CountAndLabel` call?

6. After the *first* `AddAllSafeEdges` call, how many components (at most) are there? (i.e., what is the maximum possible value of `count`?)

7. What is the worst-case number of iterations of the `while`-loop?

8. What is the running time of subsequent `CountAndLabel` calls?

9. Give an upper bound on the running time of `AddAllSafeEdges`.


## Bonus Reading

Read pp. 262-263. *Upshot:* Borůvka's Algorithm often beats its worst-case $O(E \log V)$ running time.
