---
title: "Classical MST Algorithms"
date: "March 26, 2021"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Announcement: CS Candidate

## Invitation for Feedback

Dr. Wonjun Lee will be interviewing next week for our open CS professor position.

- **Colloquium:** MG Tent #1: 3:30-5pm, Tuesday, March 30 (location tentative).
- **Teaching Demo:** Monday, March 29, 9:10-10:15: https://youtu.be/IFvpuv5k_ZU

I have to attend the Teaching Demo, so **class is cancelled for Monday, 3/29.** (Also office hours will end early on Tuesday.)

- Please consider **attending Tuesday's colloquium** as a substitute.

# Assignment Comments, Sample Solutions

## One pass: High level description

(See Jamboard)

- Mark the edges to contract (find min, should be $O(E)$).
- For each marked edge $uv$:
    - Make a new vertex/linked list $n$
    - Add all of $u$'s neighbors to $n$, and weight them appropriately.
    - Add all of $v$'s neighbors to $n$, and weight them. (avoiding duplicates somehow)
    - Add $n$ to the linked lists of the vertices in $n$'s list.
    - Remove $u$ and $v$'s lists.

Need to add some details to this to specify completely. $O(V+E)$?

## Partial solution? Could we speed it up?

```
Boruvka(G) {
  initalize set edges_to_collapse
  
  for each vertex v in G {
      min_edge = null
      min_weight = infinity
      for each edge uv according to the adjacency list {
          if uv.weight < min_weight {
              min_edge = uv
              min_weight = uv.weight
          }
      }
      add min_edge to edges_to_collapse
  }
  
  for each edge uv in edges_to_collapse {
      add new node x to adjacency list
      for each neighbor n of u
        add n to x's linked list
        change "u" in n's linked list to "x"
      for each neighbor n of v
        if n is already in x's linked list
          if nv.weight > nu.weight
            nx.weight = nv.weight
        else
          add n to x's linked list
          change "v" in n's linked list to "x"
      remove u and v nodes from adjacency list
  }
}
```

## Another nice try

```
BoruvkaContraction(A[1..V]):
  //iterate over all edges and mark those to be contracted
  for i <- 1..V:
    edges <- A[i] //has indices 1..n
    min <- (edge with weight of infinity)
    for j <- 1..n:
      if edges[j].weight < min.weight:
        min <- the edge from i to edges[j]
    mark the edge min
  for i <- 1..V:
    if A[i] contains a marked edge uv:
      //add v's out-edges to u
      new <- A[u]
      old <- A[v]
      for j <- 1..n:
        if old[j] is not in new:
          add old to A[u] // new
        else: //only add smallest parallel edge
          parallel <- from new to old[j]
          if old[j].weight < parallel.weight:
            delete parallel from A[u] // new
            add A[v][j] to A[u]
        replace v in A[old[j]] with u
      delete old
```

# Classical MST Algorithms

## Prim's Theorem; Meta-algorithm

**Theorem.** Let $F$ be an intermediate spanning forest of a weighted graph $G$, and suppose $e$ is a safe edge. Then the minimum spanning tree contains $e$.

```{javascript, eval=FALSE}
MetaMST(V,E)
    F = (V, ∅)
    while F is not a tree
        add a safe edge to F   // NOTE: This choice needs to be specified!!
    return F
```

## Jarník's Algorithm (aka Prim's algorithm)

- Start with any vertex $T$.
- Add $T$'s safe edge to $T$.
    - Repeat until you run out of vertices.
    
Exercise: Describe this completely, and verify its running time.


## Kruskal's Algorithm

- Start with the forest $F$ of vertices.
- Scan all edges by increasing weight; if an edge is safe, add it to $F$.

```
Kruskal(V, E):
    sort E by increasing weight (e.g., MergeSort)
    F <- (V, ∅)
    for each vertex v ∈ V
        MakeSet(v) // Set data structure partitioning V
    for i <- 1 to |E|
        uv <- ith lightest edge in E
        if Find(u) != Find(v) // Are u and v in different partitions?
            Union(u, v)
            add uv to F
    return F
```

- Need $O(E \log E)$ to sort, and that's the most expensive step. 
- In general, $E$ is going to be bigger than $V$ for connected graphs, but at most $E = O(V^2)$. So $\Theta(\log E) = \Theta(\log V)$.
- So the time is $O(E \log V)$ (which is the same as $O(E \log E)$, in this case).

## Table Groups

```{r, echo=FALSE}
library(knitr)
roster <- c("Ethan", "Talia", "Drake", "Jack", "Andrew", "Blake", "Jordan", "Graham", "Kevin", "Logan", "Claire", "Bri", "Trevor", "James", "Kristen", "Levi", "Grace", "John", "Isaac", "Josiah", "Nathan")
set.seed(3262021) 
n <- length(roster)
ngps <- 7
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

## Step through Jarník and Kruskal

1. On the Jamboard, step through Jarník's algorithm on the given graph. Label the edges in the order they get added. Start at the vertex at the top.

2. On the Jamboard, step through Kruskal's algorithm on the given graph. Label the edges in the order they get added. 

# Adjacency Matrices

## Recall: Adjacency Matrix

Weighted graphs can be represented with $V \times V$ *matrix*:

$$
\begin{bmatrix}
a_{1,1} & a_{1,2} & \cdots & a_{1,V} \\
a_{2,1} & a_{2,2} & \cdots & a_{2,V} \\
\vdots & \vdots && \vdots \\
a_{V,1} & a_{V,2} & \cdots & a_{V,V} 
\end{bmatrix}
$$

- $a_{i,j} = w$ if there is an edge from vertex $i$ to vertex $j$ of weight $w$.
- $a_{i,j} = 0$ if there isn't an edge.
- More efficient for *dense* graphs (lots of edges). Less efficient for *sparse* graphs.

## Compute the adjacency matrix

3. Compute the adjacency matrix for the graph given on the Jamboard.


